---
title: "Sémiologie Graphique — Bertin 1967"
subtitle: "10 représentations ggplot2 des 90 départements français"
output: html_notebook
---

Jacques Bertin's *Sémiologie graphique* (1967) defined the grammar of graphics through **retinal variables**: position, size, value (lightness), texture, colour, orientation, and shape. His canonical dataset — the active population of 90 French *départements* split across agriculture, industry, and services — illustrates how the choice and ordering of these variables reveal structure.

```{r package_setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(here)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 11,
  fig.height = 7
)

sector_colors <- c(Agriculture = "#4daf4a", Industry = "#ff7f00", Services = "#377eb8")

read_csv_locale <- function(path, ...) {
  readr::read_csv(
    path,
    locale = readr::locale(decimal_mark = ".", grouping_mark = ","),
    ...
  )
}
```

```{r load_data}
bertin <- read_csv_locale(here::here("course", "bertin-1967.csv")) |>
  mutate(
    dominant = case_when(
      agriculture >= industry & agriculture >= services ~ "Agriculture",
      industry   >= agriculture & industry >= services ~ "Industry",
      TRUE                                             ~ "Services"
    ),
    dominant = factor(dominant, levels = c("Agriculture", "Industry", "Services"))
  )

bertin_long <- bertin |>
  pivot_longer(
    cols      = c(agriculture, industry, services),
    names_to  = "sector",
    values_to = "pct"
  ) |>
  mutate(
    sector = recode(sector,
      agriculture = "Agriculture",
      industry    = "Industry",
      services    = "Services"
    ),
    sector = factor(sector, levels = c("Agriculture", "Industry", "Services"))
  )

glimpse(bertin)
```

```{r data_table}
knitr::kable(
  head(bertin, 20),
  caption = "Bertin (1967) — 20 premiers départements (% actifs par secteur)",
  col.names = c("Code", "Département", "Agriculture", "Industrie", "Services", "Secteur dominant")
)
```

---

## 1. Lollipop chart — variable visuelle : POSITION (ordonné)

The simplest and most powerful Bertin tool: **ordering** a single quantitative variable maps it to position on a common scale, allowing immediate rank-reading.

```{r plot_01, fig.height=14, fig.width=9}
bertin |>
  arrange(agriculture) |>
  mutate(department = factor(department, levels = department)) |>
  ggplot(aes(x = agriculture, y = department)) +
  geom_segment(aes(x = 0, xend = agriculture, yend = department),
               colour = sector_colors["Agriculture"], alpha = 0.45) +
  geom_point(colour = sector_colors["Agriculture"], size = 2.5) +
  labs(
    title    = "% actifs en agriculture par département (trié)",
    subtitle = "Variable visuelle : POSITION — Bertin (1967) §2.3",
    x = "Agriculture (%)", y = NULL
  ) +
  theme_minimal(base_size = 22) +
  theme(axis.text.y = element_text(size = 7))
```

---

## 2. Matrice réordonnée — variable visuelle : VALEUR (clarté)

Bertin's signature contribution: the **reorderable matrix**. Rows (départements) are sorted by agriculture; columns are the three sectors. Cell lightness encodes magnitude. Reordering reveals the dominant gradient.

```{r plot_02, fig.height=14, fig.width=5}
dept_order <- bertin |> arrange(desc(agriculture)) |> pull(department)

bertin_long |>
  mutate(department = factor(department, levels = dept_order)) |>
  ggplot(aes(x = sector, y = department, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "white", high = "#1a3a6b", name = "%") +
  labs(
    title    = "Matrice réordonnée — population active par secteur",
    subtitle = "Variable visuelle : VALEUR — lignes triées par agriculture décroissante",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y   = element_text(size = 6),
    legend.position = "right"
  )
```

---

## 3. Barres empilées 100 % — variable visuelle : TAILLE (aire)

Area proportional to share. Sorting by agriculture makes the agricultural–industrial gradient legible along the x-axis — a direct application of Bertin's principle of **ordered series**.

```{r plot_03, fig.height=6, fig.width=13}
bertin |>
  arrange(desc(agriculture)) |>
  mutate(department = factor(department, levels = department)) |>
  pivot_longer(c(agriculture, industry, services),
               names_to = "sector", values_to = "pct") |>
  mutate(
    sector = recode(sector,
      agriculture = "Agriculture", industry = "Industry", services = "Services"),
    sector = factor(sector, levels = c("Services", "Industry", "Agriculture"))
  ) |>
  ggplot(aes(x = department, y = pct, fill = sector)) +
  geom_col(width = 1) +
  scale_fill_manual(values = sector_colors) +
  labs(
    title    = "Composition de la population active — barres empilées 100 %",
    subtitle = "Variable visuelle : TAILLE — triées par agriculture décroissante (gauche → droite)",
    x = NULL, y = "%", fill = NULL
  ) +
  theme_minimal(base_size = 22) +
  theme(
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10),
    legend.position = "top"
  )
```

---

## 4. Nuage de points — variable visuelle : POSITION (deux axes)

Two quantitative variables mapped to x and y. The strong negative correlation between agriculture and industry is the key structural finding of Bertin's analysis — urban-industrial France vs rural France.

```{r plot_04}
label_dept <- bertin |>
  filter(agriculture > 55 | industry > 50 | agriculture < 3)

bertin |>
  ggplot(aes(x = agriculture, y = industry, colour = dominant)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey50",
              linetype = "dashed", linewidth = 0.7) +
  geom_text(data = label_dept, aes(label = department),
            size = 2.8, hjust = -0.08, vjust = 0.4) +
  scale_colour_manual(values = sector_colors) +
  labs(
    title    = "Agriculture vs Industrie",
    subtitle = "Variable visuelle : POSITION — corrélation négative, couleur = secteur dominant",
    x = "Agriculture (%)", y = "Industrie (%)", colour = "Secteur dominant"
  ) +
  theme_minimal(base_size = 22)
```

---

## 5. Carte à bulles — variable visuelle : TAILLE (surface)

Three variables in one plane: position (agriculture, services) + size (industry). Bertin called size the only truly **quantitative** retinal variable after position.

```{r plot_05}
bertin |>
  ggplot(aes(x = agriculture, y = services,
             size = industry, colour = dominant)) +
  geom_point(alpha = 0.65) +
  scale_size_area(max_size = 15, name = "Industrie (%)") +
  scale_colour_manual(values = sector_colors) +
  labs(
    title    = "Agriculture × Services × Industrie (taille)",
    subtitle = "Variable visuelle : TAILLE — surface proportionnelle à % industrie",
    x = "Agriculture (%)", y = "Services (%)", colour = "Secteur dominant"
  ) +
  theme_minimal(base_size = 22)
```

---

## 6. Coordonnées parallèles — profils multivariés

Each line is one département; each vertical axis is one sector (normalised 0–1). Lines crossing from high agriculture to high services reveal the **bipolarity** Bertin identified in the data.

```{r plot_06}
scale01 <- function(x) (x - min(x)) / (max(x) - min(x))

bertin |>
  mutate(across(c(agriculture, industry, services), scale01,
                .names = "{.col}_n")) |>
  pivot_longer(cols = ends_with("_n"),
               names_to = "sector", values_to = "value") |>
  mutate(
    sector = recode(sector,
      agriculture_n = "Agriculture", industry_n = "Industry", services_n = "Services"),
    sector = factor(sector, levels = c("Agriculture", "Industry", "Services"))
  ) |>
  ggplot(aes(x = sector, y = value,
             group = department, colour = dominant)) +
  geom_line(alpha = 0.7, linewidth = 1) +
  scale_colour_manual(values = sector_colors) +
  labs(
    title    = "Coordonnées parallèles — profils des 90 départements",
    subtitle = "Variable visuelle : POSITION — chaque ligne = un département, valeurs normalisées [0–1]",
    x = NULL, y = "Valeur normalisée", colour = "Secteur dominant"
  ) +
  theme_minimal(base_size = 22)
```

---

## 7. Histogrammes — distributions marginales

How is each sector distributed across the 90 départements? Agriculture is right-skewed (most départements have moderate agriculture); industry and services show distinct modes.

```{r plot_07, fig.height=7}
bertin_long |>
  ggplot(aes(x = pct, fill = sector)) +
  geom_histogram(bins = 16, colour = "white", linewidth = 0.3) +
  facet_wrap(~sector, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = sector_colors) +
  labs(
    title    = "Distribution de la population active par secteur",
    subtitle = "Variable visuelle : POSITION — histogrammes des 90 départements",
    x = "% actifs", y = "Nombre de départements"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 8. Graphe divergent — écart à la moyenne nationale

Signed deviation from the national mean highlights **above/below-average** départements. Bertin emphasised that ordering and a reference line turn raw values into a relational reading.

```{r plot_08, fig.height=14, fig.width=9}
avg_agri <- mean(bertin$agriculture)

bertin |>
  mutate(
    dev  = agriculture - avg_agri,
    department = factor(department, levels = department[order(dev)])
  ) |>
  ggplot(aes(x = dev, y = department, fill = dev > 0)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "black", linewidth = 0.5) +
  scale_fill_manual(
    values = c("FALSE" = "#0072B2", "TRUE" = "#E69F00"),
    labels = c("Sous la moyenne", "Au-dessus de la moyenne"),
    name   = NULL
  ) +
  labs(
    title    = paste0("Écart au taux national d'agriculture (",
                      round(avg_agri, 1), " %)"),
    subtitle = "Variable visuelle : POSITION (signée) — divergence autour de la référence nationale",
    x = "Écart (%)", y = NULL
  ) +
  theme_minimal(base_size = 22) +
  theme(
    axis.text.y    = element_text(size = 7),
    legend.position = "top"
  )
```

---

## 9. Petits multiples — comparaison par secteur dominant

Faceting by dominant sector is Bertin's **selective reading**: isolate one category at a time while keeping the same visual encoding. Each panel is sorted internally by agriculture.

```{r plot_09, fig.height=10, fig.width=12}
dept_levels <- bertin |>
  arrange(dominant, desc(agriculture)) |>
  pull(department)

bertin |>
  mutate(department = factor(department, levels = dept_levels)) |>
  ggplot(aes(x = agriculture, y = department, colour = dominant)) +
  geom_point(size = 2.5) +
  geom_segment(aes(x = 0, xend = agriculture, yend = department),
               alpha = 0.3) +
  facet_wrap(~dominant, scales = "free_y", ncol = 3) +
  scale_colour_manual(values = sector_colors) +
  labs(
    title    = "Départements par secteur dominant — % agriculture",
    subtitle = "Petits multiples (POSITION) — chaque panneau trié par agriculture décroissante",
    x = "Agriculture (%)", y = NULL
  ) +
  theme_minimal(base_size = 22) +
  theme(
    legend.position  = "none",
    axis.text.y      = element_text(size = 12),
    strip.text       = element_text(face = "bold")
  )
```

---

## 10. Nuage étiqueté Agriculture × Services — frontières structurelles

The agriculture–services scatter reveals the three **structural poles** of the French economy in 1967. Bertin used this bipolar reading to argue that colour is the natural variable to encode nominal categories once quantitative position is exhausted.

```{r plot_10}
extremes <- bertin |>
  filter(
    agriculture == max(agriculture) |
    agriculture <= 2               |
    services    == max(services)   |
    services    <= 25              |
    industry    == max(industry)
  )

bertin |>
  ggplot(aes(x = agriculture, y = services, colour = dominant)) +
  geom_point(size = 3.5, alpha = 0.8) +
  ggplot2::geom_text(
    data    = extremes,
    aes(label = department),
    size    = 3,
    hjust   = -0.1,
    vjust   = 0.4,
    fontface = "italic"
  ) +
  scale_colour_manual(values = sector_colors) +
  labs(
    title    = "Agriculture × Services — pôles structurels de la France de 1967",
    subtitle = "Variable visuelle : COULEUR (nominale) + POSITION — frontières des secteurs dominants",
    x = "Agriculture (%)", y = "Services (%)",
    colour = "Secteur dominant"
  ) +
  theme_minimal()
```

---

## 11. Carte à camemberts — variable visuelle : COULEUR + TAILLE

Pie charts positioned at each département's geographic centroid. Position encodes geography; colour encodes sector; arc area encodes share. This is Bertin's **proportional symbol** applied to compositional data on a spatial support.

```{r plot_11, fig.width=12, fig.height=10}
france_map <- ggplot2::map_data("france")

dept_centroids <- data.frame(
  department = c(
    "Ain", "Aisne", "Allier", "Alpes-de-Haute-Provence", "Hautes-Alpes",
    "Alpes-Maritimes", "Ardèche", "Ardennes", "Ariège", "Aube",
    "Aude", "Aveyron", "Bouches-du-Rhône", "Calvados", "Cantal",
    "Charente", "Charente-Maritime", "Cher", "Corrèze", "Corse",
    "Côte-d'Or", "Côtes-du-Nord", "Creuse", "Dordogne", "Doubs",
    "Drôme", "Eure", "Eure-et-Loir", "Finistère", "Gard",
    "Haute-Garonne", "Gers", "Gironde", "Hérault", "Ille-et-Vilaine",
    "Indre", "Indre-et-Loire", "Isère", "Jura", "Landes",
    "Loir-et-Cher", "Loire", "Haute-Loire", "Loire-Atlantique", "Loiret",
    "Lot", "Lot-et-Garonne", "Lozère", "Maine-et-Loire", "Manche",
    "Marne", "Haute-Marne", "Mayenne", "Meurthe-et-Moselle", "Meuse",
    "Morbihan", "Moselle", "Nièvre", "Nord", "Oise",
    "Orne", "Pas-de-Calais", "Puy-de-Dôme", "Basses-Pyrénées", "Hautes-Pyrénées",
    "Pyrénées-Orientales", "Bas-Rhin", "Haut-Rhin", "Rhône", "Haute-Saône",
    "Saône-et-Loire", "Sarthe", "Savoie", "Haute-Savoie", "Seine",
    "Seine-Maritime", "Seine-et-Marne", "Seine-et-Oise", "Deux-Sèvres", "Somme",
    "Tarn", "Tarn-et-Garonne", "Var", "Vaucluse", "Vendée",
    "Vienne", "Haute-Vienne", "Vosges", "Yonne", "Territoire de Belfort"
  ),
  lon = c(
     5.33,  3.47,  3.17,  6.25,  6.27,
     7.10,  4.40,  4.73,  1.60,  4.08,
     2.50,  2.55,  5.40, -0.37,  2.57,
     0.22, -0.80,  2.52,  2.00,  9.10,
     4.83, -2.80,  2.00,  0.75,  6.43,
     5.20,  1.05,  1.30, -4.08,  4.22,
     1.40,  0.58, -0.52,  3.40, -1.70,
     1.57,  0.72,  5.57,  5.58, -0.70,
     1.38,  4.12,  3.72, -1.72,  2.00,
     1.60,  0.48,  3.50, -0.60, -1.40,
     4.10,  5.20, -0.58,  6.20,  5.30,
    -2.80,  6.70,  3.52,  3.20,  2.50,
     0.10,  2.33,  3.10, -0.80,  0.28,
     2.60,  7.48,  7.32,  4.70,  6.10,
     4.60,  0.20,  6.47,  6.42,  2.35,
     1.05,  3.00,  2.02, -0.32,  2.32,
     2.18,  1.28,  6.30,  5.10, -1.42,
     0.50,  1.30,  6.48,  3.52,  6.87
  ),
  lat = c(
    46.10, 49.52, 46.40, 44.05, 44.65,
    43.90, 44.75, 49.70, 42.90, 48.30,
    43.10, 44.28, 43.48, 49.10, 45.10,
    45.70, 45.80, 47.00, 45.30, 42.20,
    47.42, 48.50, 46.00, 45.10, 47.20,
    44.70, 49.12, 48.43, 48.22, 43.90,
    43.40, 43.68, 44.80, 43.60, 48.10,
    46.80, 47.22, 45.28, 46.78, 44.00,
    47.60, 45.52, 45.07, 47.33, 47.90,
    44.62, 44.38, 44.50, 47.40, 49.00,
    49.00, 48.08, 48.18, 48.72, 48.90,
    47.80, 49.00, 47.10, 50.52, 49.38,
    48.60, 50.50, 45.78, 43.30, 43.07,
    42.60, 48.52, 47.92, 45.78, 47.62,
    46.60, 48.00, 45.50, 46.00, 48.88,
    49.62, 48.63, 48.88, 46.52, 49.90,
    43.78, 44.08, 43.48, 44.00, 46.72,
    46.60, 45.88, 48.18, 47.80, 47.62
  ),
  stringsAsFactors = FALSE
)

bertin_geo <- merge(bertin, dept_centroids, by = "department")

make_pie_polygons <- function(data, r = 0.38) {
  sectors <- c("Agriculture", "Industry", "Services")
  out     <- vector("list", nrow(data) * 3L)
  k       <- 0L
  for (i in seq_len(nrow(data))) {
    vals   <- c(data$agriculture[i], data$industry[i], data$services[i]) / 100
    lon    <- data$lon[i]
    lat    <- data$lat[i]
    dept   <- data$department[i]
    starts <- pi / 2 - c(0, cumsum(vals[-3])) * 2 * pi
    ends   <- pi / 2 - cumsum(vals) * 2 * pi
    lon_scale <- 1 / cos(lat * pi / 180)
    for (j in 1:3) {
      k        <- k + 1L
      theta    <- seq(starts[j], ends[j], length.out = 20)
      out[[k]] <- data.frame(
        x          = c(lon, lon + r * lon_scale * cos(theta), lon),
        y          = c(lat, lat + r * sin(theta), lat),
        sector     = sectors[j],
        department = dept,
        group      = paste(dept, sectors[j]),
        stringsAsFactors = FALSE
      )
    }
  }
  out        <- do.call(rbind, out)
  out$sector <- factor(out$sector, levels = c("Agriculture", "Industry", "Services"))
  out
}

pie_data <- make_pie_polygons(bertin_geo, r = 0.3)

ggplot() +
  geom_polygon(
    data  = france_map,
    aes(x = long, y = lat, group = group),
    fill    = "grey92", colour = "white", linewidth = 0.25
  ) +
  geom_polygon(
    data  = pie_data,
    aes(x = x, y = y, group = group, fill = sector),
    colour = NA
  ) +
  scale_fill_manual(values = sector_colors) +
  coord_quickmap() +
  labs(
    title    = "11. Carte à camemberts — population active par département (1967)",
    subtitle = "Variable visuelle : COULEUR + TAILLE (aire) — rayon fixe, secteurs en proportion",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    panel.grid      = element_blank(),
    legend.position = "bottom"
  )
```

---

## 12. Petits multiples : barres par département (12 colonnes)

One panel per département, three bars per panel, sorted from most agricultural (top-left) to least (bottom-right). The grid reading replicates Bertin's **ordered matrix** in bar form — local bar height, global gradient.

```{r plot_12, fig.width=18, fig.height=12}
dept_ordered <- bertin |>
  arrange(desc(agriculture)) |>
  pull(department)

bertin_long |>
  mutate(department = factor(department, levels = dept_ordered)) |>
  ggplot(aes(x = sector, y = pct, fill = sector)) +
  geom_col(width = 0.85) +
  scale_fill_manual(values = sector_colors) +
  facet_wrap(~department, ncol = 12) +
  labs(
    title    = "12. Population active par secteur — 90 départements",
    subtitle = "Petits multiples (12 colonnes) — triés par agriculture décroissante",
    x = NULL, y = "%", fill = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x     = element_blank(),
    axis.ticks.x    = element_blank(),
    strip.text      = element_text(size = 10),
    panel.spacing   = unit(0.15, "lines"),
    legend.position = "bottom"
  )
```
