---
title: "Kanban Simulation Metrics"
output:
  ioslides_presentation:
    theme: flatly
    widescreen: true
author: "Alexandre Masselot"
date: "2027-02-18"
---
```{r package_setup, echo = FALSE, message = FALSE, warning = FALSE}
# https://chatgpt.com/c/699f73a9-eb58-8337-9f49-6b0525b21ff7
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
  # Interactive in RStudio: use the current editor document path
  rmd_path <- rstudioapi::getActiveDocumentContext()$path
} else {
  # When knitting/rendering: knitr knows the input
  rmd_path <- knitr::current_input(dir = TRUE)
}
if (!is.null(rmd_path) && nzchar(rmd_path)) {
  knitr::opts_knit$set(root.dir = dirname(normalizePath(rmd_path)))
  setwd(dirname(normalizePath(rmd_path)))
}

knitr::opts_chunk$set(
  fig.width = 12,   # increase width
  fig.height = 6,
  out.width = "100%"   # VERY IMPORTANT
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

```
```{r loading_data}
df.daily.status = read.csv("kanban-daily-by-status.csv")

#convert $date to date
df.daily.status$date = as.Date(df.daily.status$date)

# order the statuses
status_list = c("BACKLOG", "IN_SCOPING", "DONE_SCOPING", "IN_DEVELOPMENT", "DONE_DEVELOPMENT", "IN_TESTING", "DONE_TESTING", "DEPLOYED")
df.daily.status$status <- factor(
  df.daily.status$status,
  levels = status_list
)

#assign status => colors
status_colors = c(
  "BACKLOG" = rgb(231, 41, 138, alpha = 100, maxColorValue = 255),
  "IN_SCOPING" = rgb(231, 41, 138, maxColorValue = 255),
  "DONE_SCOPING" = rgb(117, 112, 179, alpha = 100, maxColorValue = 255),
  "IN_DEVELOPMENT" = rgb(117, 112, 179, maxColorValue = 255),
  "DONE_DEVELOPMENT" = rgb(217, 95, 2, alpha = 100, maxColorValue = 255),
  "IN_TESTING" = rgb(217, 95, 2, maxColorValue = 255),
  "DONE_TESTING" = rgb(27, 158, 119, alpha = 100, maxColorValue = 255),
  "DEPLOYED" = rgb(27, 158, 119, maxColorValue = 255)
)

# fill missing values with zero
df.daily.status <- df.daily.status %>%
  complete(date, status, fill = list(nb = 0))
```
# A board to rule them all


```{r few stats, echo=FALSE}
df.by.day = df.daily.status %>%
  group_by(date) %>%
  summarise(n = sum(nb))
nb_tickets = max(df.by.day$n)
start_date = min(df.daily.status$date)
end_date = max(df.daily.status$date)
```


```{r demo_data}
df.demo = data.frame(
  status = c(status_list, status_list),
  date = c(rep("x", length(status_list)), rep("x + 1", length(status_list))),
  n = c(c(5, 2, 3, 4, 3, 2, 1, 6), c(6, 1, 4, 2, 3, 4, 1, 7))
) %>%
  mutate(status = factor(status, levels = status_list))

df.demo.push = data.frame(
  status = c("IN_SCOPING", "IN_DEVELOPMENT", "IN_TESTING", "DEPLOYED"),
  date = rep("x", 4),
  n = c(10, 7, 2, 7)
) %>%
  mutate(status = factor(status, levels = status_list))
```
```{r plot_board, echo=FALSE}
plot_board = function(df.tickets, max_tickets) {
  ggplot(df.tickets, aes(x = status, y = n, fill = status)) +
    geom_col(width = 0.9) +
    geom_hline(
      yintercept = 0:max_tickets,
      color = "white",
      linewidth = 3
    ) +
    scale_y_reverse(expand = c(0, 0)) +
    scale_x_discrete(position = "top", expand = expansion(mult = c(0.05, 0.2))) +
    scale_fill_manual(values = status_colors, drop = FALSE) +
    theme_minimal() +
    labs(
      title = "Status per day",
      x = "",
      y = ""
    ) +
    theme(
      panel.grid = element_blank(),   # remove default grid so only your white lines remain
      legend.position = "none",
      axis.text.x = element_text(
        size = 12,
        face = "bold",
        angle = 30,
        hjust = 0,   # adjust horizontal alignment
        vjust = 0    # adjust vertical alignment
      ),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}
```
## The board

A kanban board represents the status of work at a given moment in time.

Tickets move from left to right.

```{r plot_demo_push}
plot_board(df.demo.push, max(df.demo.push$n))
```

## Kanban is pulling, not pushing!

Greyed colors represent ticket in waiting stages.

```{r board_demo}
df.demo = data.frame(
  status = c(status_list, status_list),
  date = c(rep("x", length(status_list)), rep("x + 1", length(status_list))),
  n = c(c(5, 2, 3, 4, 3, 2, 1, 6), c(6, 1, 4, 2, 3, 4, 1, 7))
) %>%
  mutate(status = factor(status, levels = status_list))

max_tickets = max(df.demo$n)


plot_board(subset(df.demo, date == "x"), max_tickets)
```

## Day + 1 evolution

The board evolves every day.
```{r demo_board_two_days}
plot_board(df.demo, max_tickets) +
  facet_wrap(~date, nrow = 1, strip.position = "bottom") +
  labs(
    y = "nb tickets"
  ) +
  theme(
    strip.text.x = element_text(
      size = 16,
      face = "bold",
    )
  )
```

## Daily stacked

We can also see a day as a stacked histogram (no ticket details).
```{r demo_daily_stacked}

plot_daily_stack <- function(df.tickets) {
  ggplot(df.tickets, aes(x = date, y = n, fill = status)) +
    geom_col(width = 0.7) +
    scale_fill_manual(values = status_colors) +
    labs(
      y = "nb tickets"
    ) +
    theme_minimal(base_size = 22) +
    theme(
      panel.grid.major.x = element_blank(),
      legend.position = "right",
      legend.text = element_text(size = 14)
    )
}
plot_daily_stack(subset(df.demo, date == "x"))
```


## Stack evolution

```{r demo_daily_stacked_two_days}
plot_daily_stack(df.demo)
```

# Let's go wider

## Larger simulation
With `simulate_kanban.py` a ticket evolution was simulated with `r nb_tickets` tickets, starting on `r start_date` and ending on `r end_date`.

WIP limit and velocity are changing over time due to hiring and other tuning factors.

## Cumulative flow diagram

```{r plot_cfd_function, echo = FALSE}
kanban_guide_axis_x <- function() {
  list(
    coord_cartesian(xlim = c(start_date, end_date)),
    scale_x_date(date_labels = "%y/%m")
  )
}

plot_cfd = function() {
  p = ggplot(
    df.daily.status,
    aes(x = date, y = nb, group = status, fill = status)
  ) +
    geom_area(position = "stack") +
    kanban_guide_axis_x() +
    scale_fill_manual(values = status_colors) +
    theme_minimal(base_size = 22) +
    theme(
      legend.text = element_text(size = 14)
    ) +
    labs(
      title = "Cumulative Flow Diagram",
      x = "time",
      y = "nb tickets"
    )
  p
}
```

```{r cfd_zoom}

plot_cfd() +
  coord_cartesian(xlim = c(as.Date("2026-01-31"), as.Date("2026-04-15")), ylim = c(0, 110)) +
  annotate(
    "segment",
    x = as.Date("2026-01-31"),
    xend = as.Date("2026-04-01"),
    y = 24,
    yend = 24,
    lwd = 1,
    arrow = arrow(
      length = unit(0.4, "cm"),
      type = "closed",
      ends = "both"
    )
  ) +
  annotate(
    "label",
    x = as.Date("2026-02-20"),
    y = 28,
    label = "Lead time",
    hjust = 0.5,
    vjust = 0,
    size = 6,
    padding = unit(0.5, "lines")
  ) +
  annotate(
    "segment",
    x = as.Date("2026-03-20"),
    xend = as.Date("2026-03-20"),
    y = 16,
    yend = 70,
    lwd = 1,
    arrow = arrow(
      length = unit(0.4, "cm"),
      type = "closed",
      ends = "both"
    )
  ) +
  annotate(
    "label",
    x = as.Date("2026-03-22"),
    y = 56,
    hjust = 0,
    label = "WIP\n(Work in Progress)",
    size = 6,
    padding = unit(0.5, "lines")
  ) +
  annotate(
    "segment",
    x = as.Date("2026-03-25"),
    xend = as.Date("2026-04-15"),
    y = 18,
    yend = 28,
    lwd = 1
  ) +
  annotate(
    "segment",
    x = as.Date("2026-03-25"),
    xend = as.Date("2026-04-15"),
    y = 18,
    yend = 18,
    lwd = 1
  ) +
  annotate(
    "label",
    x = as.Date("2026-04-05"),
    y = 14,
    hjust = 0.5,
    vjust = 1,
    label = "angle = throughput",
    size = 6,
    padding = unit(0.5, "lines")
  )
```

## CFD: the full picture

```{r plot_cfd_full}
df.annotations = data.frame(
  x = c(as.Date("2026-04-01"), as.Date("2026-07-01"), as.Date("2026-10-01")),
  y = c(170, 260, 350),
  text = c(
    "Development team doubled\n(WIP limit increases to 6)",
    "We enhance developers with AI\nvelocity is doubled",
    "More automated testing in dev (time + 50%)\nTesting time divided by 2"
  )
)

plot_cfd() +
  geom_label(
    data = df.annotations,
    aes(x = x, y = y, label = text),
    vjust = 0,
    hjust = 0.5,
    size = 3.5,
    color = "black",
    fill = "white",      # background color
    label.size = 0.3,
    label.padding = unit(0.5, "lines"),
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = df.annotations,
    aes(x = x, xend = x, y = y, yend = y - 50),
    arrow = arrow(length = unit(0.25, "cm")),
    color = "black",
    inherit.aes = FALSE
  )

```

# Throughput

```{r data_status}
df.ticket.status = read.csv("kanban-ticket-status-transitions.csv")
df.ticket.status = df.ticket.status %>%
  mutate(across(-ticket_id, as.Date))

df.deployed.per.day = df.ticket.status %>%
  filter((DEPLOYED < max(DEPLOYED, na.rm = TRUE) - 45) & (DEPLOYED > min(DEPLOYED, na.rm = TRUE) + 20)) %>%
  count(DEPLOYED, name = "n") %>%
  rename(date = DEPLOYED)
```

## Life is Noisy


```{r plot_deployed_per_day_bar_plot}
ggplot(
  data = df.deployed.per.day,
  mapping = aes(x = date, y = n)
) +
  geom_bar(stat = "identity") +
  coord_cartesian(
    xlim = c(start_date, end_date),
    ylim = c(0, max(df.deployed.per.day$n))
  ) +
  scale_x_date(date_labels = "%y/%m") +
  labs(
    title = "Ticket throughput",
    x = "date (deployment)",
    y = "deployed / week"
  ) +
  theme_minimal(base_size = 22) +
  kanban_guide_axis_x()
```

## Averaging throughput

```{r plot_deployed_per_day_points}
plot_throughput = function(hide_loess = FALSE) {
  p = ggplot(df.deployed.per.day, aes(x = date, y = n)) +
    geom_point(size = 2, alpha = 0.5)
  if (!hide_loess) {
    p <- p + geom_smooth(
      method = "loess",
      se = TRUE,
      lwd = 2
    )
  }
  p +
  coord_cartesian(
    xlim = c(start_date, end_date),
    ylim = c(0, max(df.deployed.per.day$n))
  ) +
  scale_x_date(date_labels = "%y/%m") +
    labs(
      title = "Ticket throughput",
      x = "date (deployment)",
      y = "deployed / week"
    ) +
    theme_minimal(base_size = 22)
}
plot_throughput(hide_loess = TRUE)
```


## Averaging throughput

```{r plot_deployed_per_day_loess}
plot_throughput()
```

# Lead/Cycle times

```{r data_lead_time}
df.ticket.status$lead_time = df.ticket.status$DEPLOYED - df.ticket.status$BACKLOG
df.ticket.status$cycle_time = df.ticket.status$DEPLOYED - df.ticket.status$IN_DEVELOPMENT
```

## Lead time evolution

Lead time: from backlog to deployment.

```{r plot_lead_time_line_plot}
plot_lead_time = function() {
  ggplot(df.ticket.status, aes(x = BACKLOG, y = lead_time)) +
    geom_point(size = 2, alpha = 0.5) +
    geom_smooth(
      method = "loess",
      se = TRUE,
      lwd = 2
    ) +
    kanban_guide_axis_x() +
    labs(
      title = "Lead time evolution",
      x = "date (in backlog)",
      y = "nb days"
    ) +
    theme_minimal(base_size = 22)
}
plot_lead_time()
```

## Cycle time evolution

Cycle time: from development taking over to deployment.

```{r plot_cycle_time_line_plot}
plot_cycle_time = function() {
  ggplot(df.ticket.status, aes(x = IN_DEVELOPMENT, y = cycle_time)) +
    geom_point(size = 2, alpha = 0.5) +
    geom_smooth(
      method = "loess",
      se = TRUE,
      lwd = 2
    ) +
    kanban_guide_axis_x() +
    labs(
      title = "Cycle time evolution",
      x = "date (in dev)",
      y = "nb days"
    ) +
    theme_minimal(base_size = 22)
}
plot_cycle_time()
```

## Recap

```{r recap}
(plot_cfd() +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") | plot_throughput() + theme_minimal(base_size = 14)) /
  (plot_lead_time() + theme_minimal(base_size = 14) | plot_cycle_time() + theme_minimal(base_size = 14))
```


