---
title: "Geneva Tree Inventory: Exploring Urban Canopy with R and ggplot2"
subtitle: "A grammar-of-graphics journey through the cantonal isolated tree inventory"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
    df-print: paged
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Bootstrap and Data Acquisition {#story-01}

This section sets up the project environment, locates or downloads the cantonal tree inventory, and converts it from a Shapefile bundle into a single GeoPackage file. The **learning objective** is to understand how geospatial data reaches R and how format choices affect long-term reproducibility. The **grammar-of-graphics concept** introduced here is that every visualisation begins with a clean, well-understood data source — garbage in, garbage out applies to maps as much as to bar charts.

```{r packages}
library(here)
library(sf)
library(tidyverse)
library(naniar)
library(patchwork)
library(scales)
library(viridis)
library(treemapify)
library(ggrepel)
library(broom)
library(ggspatial)
library(geodata)
library(colorspace)
library(lubridate)

theme_set(theme_minimal(base_size = 12))

tibble(
  package = c("sf", "tidyverse", "patchwork", "viridis", "treemapify",
              "ggrepel", "broom", "ggspatial", "geodata", "naniar",
              "colorspace", "lubridate"),
  version = map_chr(package, \(p) as.character(packageVersion(p)))
) |> print(n = Inf)
```

```{r constants}
# Update if SITG changes download locations
SITG_TREE_URL    <- "https://ge.ch/sitg/geodata/SITG/OPENDATA/SIPV_ICA_ARBRE_ISOLE-SHP.zip"
SITG_COMMUNE_URL <- "https://ge.ch/sitg/geodata/SITG/OPENDATA/ARIG_COMMUNE-SHP.zip"
```

```{r acquire-data}
dir.create(here("data", "raw"),       recursive = TRUE, showWarnings = FALSE)
dir.create(here("data", "processed"), recursive = TRUE, showWarnings = FALSE)

# Accept data from the bundled location (pre-downloaded) or the standard raw location
shp_bundle <- here("data", "SIPV_ICA_ARBRE_ISOLE-SHP", "SIPV_ICA_ARBRE_ISOLE.shp")
shp_raw    <- here("data", "raw", "SIPV_ICA_ARBRE_ISOLE.shp")

shp_path <- if (file.exists(shp_bundle)) {
  shp_bundle
} else if (file.exists(shp_raw)) {
  shp_raw
} else {
  message("Downloading tree inventory from SITG …")
  tmp <- tempfile(fileext = ".zip")
  download.file(SITG_TREE_URL, tmp, mode = "wb", quiet = TRUE)
  utils::unzip(tmp, exdir = here("data", "raw"))
  list.files(here("data", "raw"), pattern = "\\.shp$", recursive = TRUE, full.names = TRUE)[1]
}

message("Shapefile located: ", shp_path)

# Verify the four canonical sibling files exist
for (ext in c(".shp", ".shx", ".dbf", ".prj")) {
  f <- sub("\\.shp$", ext, shp_path)
  stopifnot("Shapefile sibling missing" = file.exists(f))
}
```

```{r read-and-convert}
trees_raw <- sf::st_read(shp_path, quiet = FALSE)

gpkg_path <- here("data", "processed", "trees_geneva.gpkg")
sf::st_write(trees_raw, gpkg_path, delete_dsn = TRUE, quiet = TRUE)
message("GeoPackage written to: ", gpkg_path)
```

### Why four files become one

A Shapefile is not a single file — it is a bundle of at least four siblings that must travel together. The `.shp` file holds the actual geometries (point coordinates in our case); `.shx` is a spatial index that allows software to jump directly to any record without scanning the entire geometry file; `.dbf` is a legacy dBASE-III table that stores the attribute columns; and `.prj` is a plain-text description of the coordinate reference system. Lose any one of them and the others become unusable.

Converting immediately to a GeoPackage (`.gpkg`) solves all of these fragility points: one file, one format, UTF-8 character encoding by default, no 10-character field-name truncation (a Shapefile limitation inherited from the dBASE era), and a SQLite engine that can hold multiple layers in the same container. A GeoPackage is also the format recommended by the Open Geospatial Consortium for archiving and data exchange. For teaching and long-term reproducibility, it is almost always the right choice over the legacy Shapefile.

```{r verify-01}
stopifnot(
  "GeoPackage not created"        = file.exists(gpkg_path),
  "Feature count below 200 000"   = nrow(trees_raw) > 200000
)
message("Story 01 checks passed. Features: ", nrow(trees_raw))
```

With the data safely stored as a single GeoPackage, the next section examines its structure, completeness, and the accuracy limits we must keep in mind before drawing a single chart.

---

## Data Hygiene and the Honesty of Provenance {#story-02}

Before drawing a single chart, we must interrogate the data itself. This section introduces the concept of **provenance**: where did these measurements come from, who made them, and how precise are they? The grammar-of-graphics concept at work is that *every aesthetic mapping is a claim* — and the strength of that claim depends entirely on the quality of the underlying data.

```{r load-gpkg}
trees_raw <- sf::st_read(here("data", "processed", "trees_geneva.gpkg"), quiet = TRUE)
```

```{r rename-standardise}
# Rename fields to readable snake_case names that match user-story conventions.
# Key mappings (SITG field → notebook name):
#   NOM_COMPLE  → genre_espece  (species binomial, up to 50 chars)
#   HAUTEUR_TO  → hauteur       (total height, metres)
#   CIRCONFERE  → circonfere    (trunk circumference, cm; DBH = circonfere/pi)
#   DIAMETRE_C  → diametre_couronne (crown diameter, metres)
#   RAYON_COUR  → rayon_couronne (crown radius, metres)
#   DATE_PLANT  → date_plantation_raw (string YYYYMMDD)
trees <- trees_raw |>
  rename_with(tolower) |>
  rename(
    genre_espece       = nom_comple,
    hauteur            = hauteur_to,
    hauteur_tronc      = hauteur_tr,
    diametre_couronne  = diametre_c,
    rayon_couronne     = rayon_cour,
    date_plantation_raw = date_plant,
    type_plantation    = type_plant,
    stade              = stade_deve,
    esperance_vie      = esperance_
  ) |>
  mutate(
    # DBH (diameter at breast height) derived from circumference
    diametre_tronc = circonfere / pi,
    # Parse planting date: stored as YYYYMMDD string
    date_plantation = suppressWarnings(
      lubridate::ymd(date_plantation_raw)
    ),
    annee_plantation = lubridate::year(date_plantation)
  )

sf::st_write(trees, here("data", "processed", "trees_clean.gpkg"),
             delete_dsn = TRUE, quiet = TRUE)
message("Cleaned layer written: ", nrow(trees), " features")
```

```{r glimpse}
dplyr::glimpse(trees)
```

```{r summary-table}
# Tidy summary: type, non-NA count, sample value
summary_tbl <- trees |>
  sf::st_drop_geometry() |>
  select(genre_espece, hauteur, diametre_tronc, diametre_couronne,
         date_plantation, classe, vitalite, situation) |>
  summarise(across(everything(), list(
    type    = \(x) class(x)[1],
    n_valid = \(x) sum(!is.na(x)),
    pct_na  = \(x) round(100 * mean(is.na(x)), 1),
    example = \(x) as.character(x[!is.na(x)][1])
  ))) |>
  pivot_longer(everything(),
               names_to  = c("column", ".value"),
               names_sep = "_(?=[^_]+$)") |>
  arrange(pct_na |> as.numeric() |> desc())

print(summary_tbl)
```

*What to look at:* The `pct_na` column immediately reveals which fields are sparse. `hauteur` and `diametre_couronne` have some missingness; `date_plantation` is missing for a substantial share of records — we will return to that story in section 8.

```{r missingness-plot}
trees |>
  sf::st_drop_geometry() |>
  select(genre_espece, hauteur, diametre_tronc, diametre_couronne,
         date_plantation, classe, vitalite, situation, type_plantation) |>
  naniar::vis_miss(warn_large_df = FALSE) +
  labs(
    title    = "Missingness across key columns",
    subtitle = "Each column-row cell is coloured if the value is absent",
    caption  = "Source: SITG, SIPV_ICA_ARBRE_ISOLE"
  )
```

*What to look at:* Most geometric and biological fields are complete. The planting-date column stands out as systematically sparse — this is not random dropout but a structured gap driven by survey era, as the next paragraph explains.

```{r numeric-histograms}
p_haut <- ggplot(trees |> filter(!is.na(hauteur), hauteur > 0),
                 aes(x = hauteur)) +
  geom_histogram(binwidth = 2, fill = viridis::viridis(1, begin = 0.4),
                 colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(title = "Tree height distribution", x = "Total height (m)", y = "Count",
       caption = "Source: SITG, ICA")

p_circ <- ggplot(trees |> filter(!is.na(circonfere), circonfere > 0),
                 aes(x = circonfere)) +
  geom_histogram(binwidth = 20, fill = viridis::viridis(1, begin = 0.6),
                 colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " cm")) +
  labs(title = "Trunk circumference distribution", x = "Circumference (cm)", y = "Count",
       caption = "Source: SITG, ICA")

p_cour <- ggplot(trees |> filter(!is.na(diametre_couronne), diametre_couronne > 0),
                 aes(x = diametre_couronne)) +
  geom_histogram(binwidth = 1, fill = viridis::viridis(1, begin = 0.8),
                 colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(title = "Crown diameter distribution", x = "Crown diameter (m)", y = "Count",
       caption = "Source: SITG, ICA")

p_haut / p_circ / p_cour
```

*What to look at:* All three distributions are right-skewed. Most trees are small and young; a long right tail of tall or wide-crowned specimens drives the mean far above the median. This asymmetry will motivate the log-scale comparison in section 3.

### Positional accuracy: two generations of survey

::: callout-important
**Positional accuracy caveat.** The ICA dataset contains trees surveyed across several decades. Records from the 1976 baseline inventory carry a positional accuracy of approximately 25 m; records from modern digital surveys (post-2000) are accurate to approximately 1 m. This means any map-based analysis that treats all points equally risks misleading clusters and gaps, particularly in areas densely surveyed in 1976.
:::

```{r survey-decade-bar}
trees |>
  sf::st_drop_geometry() |>
  mutate(decade_obs = floor(as.numeric(substr(date_obser, 1, 4)) / 10) * 10) |>
  filter(!is.na(decade_obs)) |>
  count(decade_obs) |>
  ggplot(aes(x = factor(decade_obs), y = n)) +
  geom_col(fill = viridis::viridis(1, begin = 0.5)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Records by observation decade",
    subtitle = "Proxy for when each tree was surveyed (not when planted)",
    x        = "Observation decade",
    y        = "Number of records",
    caption  = "Source: SITG, ICA — date_obser field"
  )
```

*What to look at:* The overwhelming majority of observations cluster in the 2020s, reflecting the ongoing SEVE digital survey programme. A smaller cohort of earlier records represents trees that pre-date the modern system and may carry the higher 25 m positional uncertainty.

::: callout-note
**What this means for our charts**

1. **Density maps** (section 7) will show sharper spatial patterns for recently surveyed trees; older records may appear at slightly wrong positions, creating artefacts.
2. **Species comparisons** (section 4) may inadvertently reflect which species were surveyed recently rather than which are truly most common.
3. **Planting-date timelines** (section 8) must be interpreted alongside their missingness rate — the apparent dip in plantings before 2000 partly reflects records where the date was never digitised.
:::

```{r verify-02}
trees_clean <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
stopifnot(
  "trees_clean.gpkg missing"           = file.exists(here("data", "processed", "trees_clean.gpkg")),
  "hauteur column absent"              = "hauteur" %in% names(trees_clean),
  "diametre_tronc column absent"       = "diametre_tronc" %in% names(trees_clean),
  "genre_espece column absent"         = "genre_espece" %in% names(trees_clean)
)
message("Story 02 checks passed.")
```

Now that we know what the data can and cannot tell us, the next section builds visualisations step by step — starting with the simplest possible chart and adding one layer at a time.

---
