---
title: "Geneva Tree Inventory: Exploring Urban Canopy with R and ggplot2"
subtitle: "A grammar-of-graphics journey through the cantonal isolated tree inventory"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
    df-print: paged
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Bootstrap and Data Acquisition {#story-01}

This section sets up the project environment, locates or downloads the cantonal tree inventory, and converts it from a Shapefile bundle into a single GeoPackage file. The **learning objective** is to understand how geospatial data reaches R and how format choices affect long-term reproducibility. The **grammar-of-graphics concept** introduced here is that every visualisation begins with a clean, well-understood data source — garbage in, garbage out applies to maps as much as to bar charts.

```{r packages}
library(here)
library(sf)
library(tidyverse)
library(naniar)
library(patchwork)
library(scales)
library(viridis)
library(treemapify)
library(ggrepel)
library(broom)
library(ggspatial)
library(geodata)
library(colorspace)
library(lubridate)

theme_set(theme_minimal(base_size = 12))

tibble(
  package = c("sf", "tidyverse", "patchwork", "viridis", "treemapify",
              "ggrepel", "broom", "ggspatial", "geodata", "naniar",
              "colorspace", "lubridate"),
  version = map_chr(package, \(p) as.character(packageVersion(p)))
) |> print(n = Inf)
```

```{r constants}
# Update if SITG changes download locations
SITG_TREE_URL    <- "https://ge.ch/sitg/geodata/SITG/OPENDATA/SIPV_ICA_ARBRE_ISOLE-SHP.zip"
SITG_COMMUNE_URL <- "https://ge.ch/sitg/geodata/SITG/OPENDATA/ARIG_COMMUNE-SHP.zip"
```

```{r acquire-data}
dir.create(here("data", "raw"),       recursive = TRUE, showWarnings = FALSE)
dir.create(here("data", "processed"), recursive = TRUE, showWarnings = FALSE)

# Accept data from the bundled location (pre-downloaded) or the standard raw location
shp_bundle <- here("data", "SIPV_ICA_ARBRE_ISOLE-SHP", "SIPV_ICA_ARBRE_ISOLE.shp")
shp_raw    <- here("data", "raw", "SIPV_ICA_ARBRE_ISOLE.shp")

shp_path <- if (file.exists(shp_bundle)) {
  shp_bundle
} else if (file.exists(shp_raw)) {
  shp_raw
} else {
  message("Downloading tree inventory from SITG …")
  tmp <- tempfile(fileext = ".zip")
  download.file(SITG_TREE_URL, tmp, mode = "wb", quiet = TRUE)
  utils::unzip(tmp, exdir = here("data", "raw"))
  list.files(here("data", "raw"), pattern = "\\.shp$", recursive = TRUE, full.names = TRUE)[1]
}

message("Shapefile located: ", shp_path)

# Verify the four canonical sibling files exist
for (ext in c(".shp", ".shx", ".dbf", ".prj")) {
  f <- sub("\\.shp$", ext, shp_path)
  stopifnot("Shapefile sibling missing" = file.exists(f))
}
```

```{r read-and-convert}
trees_raw <- sf::st_read(shp_path, quiet = FALSE)

gpkg_path <- here("data", "processed", "trees_geneva.gpkg")
sf::st_write(trees_raw, gpkg_path, delete_dsn = TRUE, quiet = TRUE)
message("GeoPackage written to: ", gpkg_path)
```

### Why four files become one

A Shapefile is not a single file — it is a bundle of at least four siblings that must travel together. The `.shp` file holds the actual geometries (point coordinates in our case); `.shx` is a spatial index that allows software to jump directly to any record without scanning the entire geometry file; `.dbf` is a legacy dBASE-III table that stores the attribute columns; and `.prj` is a plain-text description of the coordinate reference system. Lose any one of them and the others become unusable.

Converting immediately to a GeoPackage (`.gpkg`) solves all of these fragility points: one file, one format, UTF-8 character encoding by default, no 10-character field-name truncation (a Shapefile limitation inherited from the dBASE era), and a SQLite engine that can hold multiple layers in the same container. A GeoPackage is also the format recommended by the Open Geospatial Consortium for archiving and data exchange. For teaching and long-term reproducibility, it is almost always the right choice over the legacy Shapefile.

```{r verify-01}
stopifnot(
  "GeoPackage not created"        = file.exists(gpkg_path),
  "Feature count below 200 000"   = nrow(trees_raw) > 200000
)
message("Story 01 checks passed. Features: ", nrow(trees_raw))
```

With the data safely stored as a single GeoPackage, the next section examines its structure, completeness, and the accuracy limits we must keep in mind before drawing a single chart.

---
