---
title: "Geneva Tree Inventory: Exploring Urban Canopy with R and ggplot2"
subtitle: "A grammar-of-graphics journey through the cantonal isolated tree inventory"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
    df-print: paged
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Bootstrap and Data Acquisition {#story-01}

This section sets up the project environment, locates or downloads the cantonal tree inventory, and converts it from a Shapefile bundle into a single GeoPackage file. The **learning objective** is to understand how geospatial data reaches R and how format choices affect long-term reproducibility. The **grammar-of-graphics concept** introduced here is that every visualisation begins with a clean, well-understood data source — garbage in, garbage out applies to maps as much as to bar charts.

```{r packages}
library(here)
library(sf)
library(tidyverse)
library(naniar)
library(patchwork)
library(scales)
library(viridis)
library(treemapify)
library(ggrepel)
library(broom)
library(ggspatial)
library(geodata)
library(colorspace)
library(lubridate)

theme_set(theme_minimal(base_size = 12))

tibble(
  package = c("sf", "tidyverse", "patchwork", "viridis", "treemapify",
              "ggrepel", "broom", "ggspatial", "geodata", "naniar",
              "colorspace", "lubridate"),
  version = map_chr(package, \(p) as.character(packageVersion(p)))
) |> print(n = Inf)
```

```{r constants}
# Update if SITG changes download locations
SITG_TREE_URL    <- "https://ge.ch/sitg/geodata/SITG/OPENDATA/SIPV_ICA_ARBRE_ISOLE-SHP.zip"
SITG_COMMUNE_URL <- "https://ge.ch/sitg/geodata/SITG/OPENDATA/ARIG_COMMUNE-SHP.zip"
```

```{r acquire-data}
dir.create(here("data", "raw"),       recursive = TRUE, showWarnings = FALSE)
dir.create(here("data", "processed"), recursive = TRUE, showWarnings = FALSE)

# Accept data from the bundled location (pre-downloaded) or the standard raw location
shp_bundle <- here("data", "SIPV_ICA_ARBRE_ISOLE-SHP", "SIPV_ICA_ARBRE_ISOLE.shp")
shp_raw    <- here("data", "raw", "SIPV_ICA_ARBRE_ISOLE.shp")

shp_path <- if (file.exists(shp_bundle)) {
  shp_bundle
} else if (file.exists(shp_raw)) {
  shp_raw
} else {
  message("Downloading tree inventory from SITG …")
  tmp <- tempfile(fileext = ".zip")
  download.file(SITG_TREE_URL, tmp, mode = "wb", quiet = TRUE)
  utils::unzip(tmp, exdir = here("data", "raw"))
  list.files(here("data", "raw"), pattern = "\\.shp$", recursive = TRUE, full.names = TRUE)[1]
}

message("Shapefile located: ", shp_path)

# Verify the four canonical sibling files exist
for (ext in c(".shp", ".shx", ".dbf", ".prj")) {
  f <- sub("\\.shp$", ext, shp_path)
  stopifnot("Shapefile sibling missing" = file.exists(f))
}
```

```{r read-and-convert}
trees_raw <- sf::st_read(shp_path, quiet = FALSE)

gpkg_path <- here("data", "processed", "trees_geneva.gpkg")
sf::st_write(trees_raw, gpkg_path, delete_dsn = TRUE, quiet = TRUE)
message("GeoPackage written to: ", gpkg_path)
```

### Why four files become one

A Shapefile is not a single file — it is a bundle of at least four siblings that must travel together. The `.shp` file holds the actual geometries (point coordinates in our case); `.shx` is a spatial index that allows software to jump directly to any record without scanning the entire geometry file; `.dbf` is a legacy dBASE-III table that stores the attribute columns; and `.prj` is a plain-text description of the coordinate reference system. Lose any one of them and the others become unusable.

Converting immediately to a GeoPackage (`.gpkg`) solves all of these fragility points: one file, one format, UTF-8 character encoding by default, no 10-character field-name truncation (a Shapefile limitation inherited from the dBASE era), and a SQLite engine that can hold multiple layers in the same container. A GeoPackage is also the format recommended by the Open Geospatial Consortium for archiving and data exchange. For teaching and long-term reproducibility, it is almost always the right choice over the legacy Shapefile.

```{r verify-01}
stopifnot(
  "GeoPackage not created"        = file.exists(gpkg_path),
  "Feature count below 200 000"   = nrow(trees_raw) > 200000
)
message("Story 01 checks passed. Features: ", nrow(trees_raw))
```

With the data safely stored as a single GeoPackage, the next section examines its structure, completeness, and the accuracy limits we must keep in mind before drawing a single chart.

---

## Data Hygiene and the Honesty of Provenance {#story-02}

Before drawing a single chart, we must interrogate the data itself. This section introduces the concept of **provenance**: where did these measurements come from, who made them, and how precise are they? The grammar-of-graphics concept at work is that *every aesthetic mapping is a claim* — and the strength of that claim depends entirely on the quality of the underlying data.

```{r load-gpkg}
trees_raw <- sf::st_read(here("data", "processed", "trees_geneva.gpkg"), quiet = TRUE)
```

```{r rename-standardise}
# Rename fields to readable snake_case names that match user-story conventions.
# Key mappings (SITG field → notebook name):
#   NOM_COMPLE  → genre_espece  (species binomial, up to 50 chars)
#   HAUTEUR_TO  → hauteur       (total height, metres)
#   CIRCONFERE  → circonfere    (trunk circumference, m; DBH = circonfere/pi)
#   DIAMETRE_C  → diametre_couronne (crown diameter, metres)
#   RAYON_COUR  → rayon_couronne (crown radius, metres)
#   DATE_PLANT  → date_plantation_raw (string YYYYMMDD)
trees <- trees_raw |>
  rename_with(tolower) |>
  rename(
    genre_espece       = nom_comple,
    hauteur            = hauteur_to,
    hauteur_tronc      = hauteur_tr,
    diametre_couronne  = diametre_c,
    rayon_couronne     = rayon_cour,
    date_plantation_raw = date_plant,
    type_plantation    = type_plant,
    stade              = stade_deve,
    esperance_vie      = esperance_
  ) |>
  mutate(
    # Convert circumference from cm to m, then derive DBH (m)
    circonfere     = circonfere / 100,
    diametre_tronc = circonfere / pi,
    # Parse planting date: stored as YYYYMMDD string
    date_plantation = suppressWarnings(
      lubridate::ymd(date_plantation_raw)
    ),
    annee_plantation = lubridate::year(date_plantation)
  )

sf::st_write(trees, here("data", "processed", "trees_clean.gpkg"),
             delete_dsn = TRUE, quiet = TRUE)
message("Cleaned layer written: ", nrow(trees), " features")
```

```{r glimpse}
dplyr::glimpse(trees)
```

```{r summary-table}
# Tidy summary: type, non-NA count, example value per column
cols_of_interest <- c("genre_espece", "hauteur", "diametre_tronc",
                      "diametre_couronne", "date_plantation",
                      "classe", "vitalite", "situation")

df_sub <- trees |> sf::st_drop_geometry() |> select(all_of(cols_of_interest))

summary_tbl <- tibble(
  column  = names(df_sub),
  type    = sapply(df_sub, \(x) class(x)[1]),
  n_valid = sapply(df_sub, \(x) sum(!is.na(x))),
  pct_na  = sapply(df_sub, \(x) round(100 * mean(is.na(x)), 1)),
  example = sapply(df_sub, \(x) as.character(x[!is.na(x)][1]))
) |>
  arrange(desc(pct_na))

print(summary_tbl)
```

*What to look at:* The `pct_na` column immediately reveals which fields are sparse. `hauteur` and `diametre_couronne` have some missingness; `date_plantation` is missing for a substantial share of records — we will return to that story in section 8.

```{r missingness-plot}
trees |>
  sf::st_drop_geometry() |>
  select(genre_espece, hauteur, diametre_tronc, diametre_couronne,
         date_plantation, classe, vitalite, situation, type_plantation) |>
  naniar::vis_miss(warn_large_data = FALSE) +
  labs(
    title    = "Missingness across key columns",
    subtitle = "Each column-row cell is coloured if the value is absent",
    caption  = "Source: SITG, SIPV_ICA_ARBRE_ISOLE"
  )
```

*What to look at:* Most geometric and biological fields are complete. The planting-date column stands out as systematically sparse — this is not random dropout but a structured gap driven by survey era, as the next paragraph explains.

```{r numeric-histograms}
p_haut <- ggplot(trees |> filter(!is.na(hauteur), hauteur > 0, hauteur < 60),
                 aes(x = hauteur)) +
  geom_histogram(binwidth = 2, fill = viridis::viridis(1, begin = 0.4),
                 colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(title = "Tree height distribution", x = "Total height (m)", y = "Count",
       caption = "Source: SITG, ICA")

p_circ <- ggplot(trees |> filter(!is.na(circonfere), circonfere > 0, circonfere < 12),
                 aes(x = circonfere)) +
  geom_histogram(binwidth = 0.2, fill = viridis::viridis(1, begin = 0.6),
                 colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(title = "Trunk circumference distribution", x = "Circumference (m)", y = "Count",
       caption = "Source: SITG, ICA")

p_cour <- ggplot(trees |> filter(!is.na(diametre_couronne), diametre_couronne > 0),
                 aes(x = diametre_couronne)) +
  geom_histogram(binwidth = 1, fill = viridis::viridis(1, begin = 0.8),
                 colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(title = "Crown diameter distribution", x = "Crown diameter (m)", y = "Count",
       caption = "Source: SITG, ICA")

p_haut / p_circ / p_cour
```

*What to look at:* All three distributions are right-skewed. Most trees are small and young; a long right tail of tall or wide-crowned specimens drives the mean far above the median. This asymmetry will motivate the log-scale comparison in section 3.

### Positional accuracy: two generations of survey

::: callout-important
**Positional accuracy caveat.** The ICA dataset contains trees surveyed across several decades. Records from the 1976 baseline inventory carry a positional accuracy of approximately 25 m; records from modern digital surveys (post-2000) are accurate to approximately 1 m. This means any map-based analysis that treats all points equally risks misleading clusters and gaps, particularly in areas densely surveyed in 1976.
:::

```{r survey-decade-bar}
trees |>
  sf::st_drop_geometry() |>
  mutate(decade_obs = floor(as.numeric(substr(date_obser, 1, 4)) / 10) * 10) |>
  filter(!is.na(decade_obs)) |>
  count(decade_obs) |>
  ggplot(aes(x = factor(decade_obs), y = n)) +
  geom_col(fill = viridis::viridis(1, begin = 0.5)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Records by observation decade",
    subtitle = "Proxy for when each tree was surveyed (not when planted)",
    x        = "Observation decade",
    y        = "Number of records",
    caption  = "Source: SITG, ICA — date_obser field"
  )
```

*What to look at:* The overwhelming majority of observations cluster in the 2020s, reflecting the ongoing SEVE digital survey programme. A smaller cohort of earlier records represents trees that pre-date the modern system and may carry the higher 25 m positional uncertainty.

::: callout-note
**What this means for our charts**

1. **Density maps** (section 7) will show sharper spatial patterns for recently surveyed trees; older records may appear at slightly wrong positions, creating artefacts.
2. **Species comparisons** (section 4) may inadvertently reflect which species were surveyed recently rather than which are truly most common.
3. **Planting-date timelines** (section 8) must be interpreted alongside their missingness rate — the apparent dip in plantings before 2000 partly reflects records where the date was never digitised.
:::

```{r verify-02}
trees_clean <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
stopifnot(
  "trees_clean.gpkg missing"           = file.exists(here("data", "processed", "trees_clean.gpkg")),
  "hauteur column absent"              = "hauteur" %in% names(trees_clean),
  "diametre_tronc column absent"       = "diametre_tronc" %in% names(trees_clean),
  "genre_espece column absent"         = "genre_espece" %in% names(trees_clean)
)
message("Story 02 checks passed.")
```

Now that we know what the data can and cannot tell us, the next section builds visualisations step by step — starting with the simplest possible chart and adding one layer at a time.

---

## Univariate Distributions and the Grammar of a Basic Chart {#story-03}

This section builds a single chart from scratch across four incremental steps, making each layer of the grammar visible before adding the next. The **learning objective** is to internalise ggplot2's layered structure: data → aesthetic → geom → stat → scale → theme. The **grammar concept** is that every chart is an accumulation of deliberate choices, not a black-box output.

```{r s03-load}
trees <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
hauteur <- trees |>
  sf::st_drop_geometry() |>
  filter(!is.na(hauteur), hauteur > 0, hauteur < 60)
```

**Step 1 — data + aesthetic only.** We declare what goes on each axis. No geometry yet, so the plot is an empty canvas with labelled axes.

```{r s03-step1}
p <- ggplot(hauteur, aes(x = hauteur))
p
```

*What to look at:* The axis limits are already set from the data range. ggplot2 knows where the data lives; we just have not told it how to draw it yet.

**Step 2 — add a geom.** `geom_histogram()` chooses `stat_bin()` automatically and draws one bar per bin.

```{r s03-step2}
p <- p + geom_histogram(binwidth = 5, colour = "white")
p
```

*What to look at:* The default grey bars and the automatic y-axis label `count` appear. The distribution is right-skewed: most trees are between 5 and 20 m, with a long tail of tall specimens.

**Step 3 — add a scale.** We apply a viridis fill to encode density visually, and format the x-axis with units.

```{r s03-step3}
p <- ggplot(hauteur, aes(x = hauteur, fill = after_stat(count))) +
  geom_histogram(binwidth = 5, colour = "white") +
  scale_fill_viridis_c(option = "D", guide = "none") +
  scale_x_continuous(labels = label_number(suffix = " m"))
p
```

*What to look at:* The fill gradient reinforces the shape of the distribution without adding a legend, because the x position already encodes height — mapping the same variable twice would be redundant.

**Step 4 — complete the chart.** Title, subtitle, axis labels, and caption are mandatory in every publication-quality figure.

```{r s03-step4}
p <- p +
  labs(
    title    = "Most Geneva trees are between 5 and 20 metres tall",
    subtitle = "Histogram of total height — 238 882 isolated trees, canton of Geneva",
    x        = "Total height (m)",
    y        = "Number of trees",
    caption  = "Source: SITG, ICA — SIPV_ICA_ARBRE_ISOLE"
  )
p
```

### How bin width changes the story

The same data under three different bin widths tells visibly different stories. Too narrow and noise dominates; too wide and structure disappears.

```{r s03-binwidth-compare}
make_hist <- function(bw, label) {
  ggplot(hauteur, aes(x = hauteur)) +
    geom_histogram(binwidth = bw, fill = viridis::viridis(1, begin = 0.5),
                   colour = "white") +
    scale_x_continuous(labels = label_number(suffix = " m")) +
    labs(title = paste0("binwidth = ", bw, " m"), x = NULL, y = "Count",
         caption = "Source: SITG, ICA")
}

make_hist(1, "1 m") | make_hist(5, "5 m") | make_hist(20, "20 m")
```

*What to look at:* At 1 m the jagged individual bars distract from the overall shape; at 20 m the bimodal suggestion in the 5–20 m range is completely hidden; at 5 m we see both the peak and the tail clearly. Bin width is a modelling decision, not a default to accept blindly.

### Density as an alternative

`geom_density()` replaces the discretisation step with a kernel smoother. It answers "what is the probability of finding a tree at this height?" rather than "how many trees fall in this bin?".

```{r s03-density}
ggplot(hauteur, aes(x = hauteur)) +
  geom_density(fill = viridis::viridis(1, begin = 0.5), alpha = 0.7,
               colour = viridis::viridis(1, begin = 0.3)) +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(
    title    = "Probability density of tree height",
    subtitle = "Kernel density estimate with default bandwidth",
    x        = "Total height (m)",
    y        = "Density",
    caption  = "Source: SITG, ICA"
  )
```

*What to look at:* The density curve smooths over the bin-width artefacts, but it implicitly extends to negative heights — a physical impossibility. Prefer `geom_density()` when the goal is shape comparison across groups; prefer `geom_histogram()` when raw counts matter.

### Log scale for trunk circumference

Trunk circumference spans two orders of magnitude (a few centimetres for a young sapling to several metres for a veteran tree), which compresses most of the distribution against the left axis on a linear scale.

```{r s03-log-compare}
circ <- trees |>
  sf::st_drop_geometry() |>
  filter(!is.na(circonfere), circonfere > 0, circonfere < 12)

p_lin <- ggplot(circ, aes(x = circonfere)) +
  geom_histogram(bins = 60, fill = viridis::viridis(1, begin = 0.5), colour = "white") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  labs(title = "Linear scale", x = "Circumference (m)", y = "Count",
       caption = "Source: SITG, ICA")

p_log <- ggplot(circ, aes(x = circonfere)) +
  geom_histogram(bins = 60, fill = viridis::viridis(1, begin = 0.7), colour = "white") +
  scale_x_log10(labels = label_number(suffix = " m")) +
  labs(title = "Log₁₀ scale", x = "Circumference (m, log scale)", y = "Count",
       caption = "Source: SITG, ICA")

p_lin | p_log
```

*What to look at:* The log scale reveals a near-normal distribution in log space — a log-normal distribution — which is exactly what allometric growth theory predicts. This will become important in section 5 when we fit the DBH–height relationship.

```{r verify-03}
stopifnot("hauteur data empty" = nrow(hauteur) > 100000)
message("Story 03 checks passed.")
```

We have now built and varied a univariate chart layer by layer. The next section scales up to a categorical variable with 200+ levels, which forces us to choose an encoding that matches the perceptual task.

---

## Categorical Encoding for High-Cardinality Data {#story-04}

Facing 1 500+ distinct species names, our first instinct might be to plot them all. This section demonstrates why that fails and what to do instead, by comparing three encodings of the same "top species" summary. The **learning objective** is to match chart type to perceptual task. The **grammar concept** is that aesthetic mappings carry a perceptual cost: position is the most accurate channel, colour the least when used for more than ~8 categories.

```{r s04-load}
trees <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)

top20 <- trees |>
  sf::st_drop_geometry() |>
  count(genre_espece, sort = TRUE) |>
  slice_head(n = 20)

top20
```

### Ordered bar chart

```{r s04-bar}
p_bar <- top20 |>
  mutate(genre_espece = fct_reorder(genre_espece, n)) |>
  ggplot(aes(x = n, y = genre_espece)) +
  geom_col(fill = viridis::viridis(1, begin = 0.5)) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title    = "Top 20 species by tree count",
    subtitle = "Ordered bar chart — position encodes rank precisely",
    x        = "Number of trees",
    y        = NULL,
    caption  = "Source: SITG, ICA"
  )
p_bar
```

### Lollipop chart

```{r s04-lollipop}
p_lollipop <- top20 |>
  mutate(genre_espece = fct_reorder(genre_espece, n)) |>
  ggplot(aes(x = n, y = genre_espece)) +
  geom_segment(aes(x = 0, xend = n, yend = genre_espece),
               colour = viridis::viridis(1, begin = 0.4)) +
  geom_point(colour = viridis::viridis(1, begin = 0.6), size = 3) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title    = "Top 20 species — lollipop variant",
    subtitle = "Reduced ink; easier to read exact values at the dot",
    x        = "Number of trees",
    y        = NULL,
    caption  = "Source: SITG, ICA"
  )
p_lollipop
```

### Treemap of the full species list

```{r s04-treemap, fig.height=8}
all_species <- trees |>
  sf::st_drop_geometry() |>
  count(genre_espece, sort = TRUE) |>
  mutate(
    label = if_else(n > 2000, genre_espece, "")
  )

p_treemap <- ggplot(all_species, aes(area = n, fill = n, label = label)) +
  treemapify::geom_treemap() +
  treemapify::geom_treemap_text(
    colour = "white", place = "centre", reflow = TRUE, size = 9
  ) +
  scale_fill_viridis_c(option = "D", labels = label_comma(),
                       name = "Trees") +
  labs(
    title    = "All species — area encodes abundance",
    subtitle = "Treemap of the full list; only species with > 2 000 trees labelled",
    caption  = "Source: SITG, ICA"
  )
p_treemap
```

### Three charts side by side

```{r s04-patchwork, fig.height=10}
(p_bar | p_lollipop) / p_treemap +
  plot_annotation(
    title   = "Three encodings of species abundance",
    caption = "Source: SITG, ICA"
  )
```

*What to look at:* The ordered bar and lollipop are essentially equivalent for ranking — both exploit position on a common scale, which Cleveland and McGill (1984) showed is the most accurately decoded visual channel. The treemap sacrifices precise rank reading in exchange for showing the entire species distribution at once, making the long tail of rare species visible. Use the bar/lollipop when the question is "which species ranks 4th?"; use the treemap when the question is "how much of the canopy do the top 5 species control?"

### The perceptual argument: Cleveland and McGill (1984)

Cleveland and McGill ranked human ability to decode visual encodings from most to least accurate:

1. Position on a common scale (bar, lollipop)
2. Position on identical non-aligned scales (faceted dots)
3. Length (bar chart without a shared baseline)
4. Angle / slope
5. Area (treemap, bubble)
6. Volume
7. Colour saturation / hue

For a ranking task, position wins. For part-of-whole with a long tail, area (treemap) communicates the structural imbalance that a bar chart of 1 500 species cannot.

### Counter-example: the pie chart

```{r s04-pie}
top20 |>
  mutate(genre_espece = fct_reorder(genre_espece, n)) |>
  ggplot(aes(x = "", y = n, fill = genre_espece)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  scale_fill_manual(
    values = colorRampPalette(viridis::viridis(8))(20),
    guide  = guide_legend(ncol = 2)
  ) +
  labs(
    title    = "Top 20 species — pie chart (do not use for ranking)",
    subtitle = "Angle and arc length are decoded far less accurately than position",
    fill     = "Species",
    caption  = "Source: SITG, ICA"
  ) +
  theme(axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())
```

*What to look at:* Can you immediately tell which species ranks 5th vs 7th? Almost certainly not — the slices share similar angles and 20 colours exceed human colour-discrimination limits. The pie chart fails the ranking task completely. It is shown here as a deliberate counter-example, not a recommendation.

```{r verify-04}
stopifnot(
  "top20 does not have 20 rows" = nrow(top20) == 20,
  "top20 not sorted"            = top20$n[1] >= top20$n[20]
)
message("Story 04 checks passed.")
```

Ordering and position are our most powerful tools for categorical data. The next section moves from one variable to two, introducing the overplotting problem and a biologically motivated model.

---

## Bivariate Relationships, Overplotting, and an Allometric Law {#story-05}

Mapping two continuous variables to x and y is conceptually simple, but 238 000 overlapping points make the naive scatter plot unreadable. This section iterates through four remedies before fitting a biologically meaningful model. The **learning objective** is to recognise and resolve overplotting. The **grammar concept** is that `stat_*` layers transform data before drawing — `stat_binhex` and `stat_density_2d` each bin the plane differently, answering different questions about the joint distribution.

```{r s05-load}
trees <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)

biv <- trees |>
  sf::st_drop_geometry() |>
  filter(
    !is.na(hauteur), hauteur > 0, hauteur < 60,
    !is.na(diametre_tronc), diametre_tronc > 0, diametre_tronc < 3
  )
```

**Step 1 — raw scatter: the overplotting problem.**

```{r s05-scatter}
ggplot(biv, aes(x = diametre_tronc, y = hauteur)) +
  geom_point(size = 0.3) +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  scale_y_continuous(labels = label_number(suffix = " m")) +
  labs(
    title    = "DBH vs height — overplotting hides all structure",
    subtitle = paste0(comma(nrow(biv)), " points fully occlude each other"),
    x        = "Trunk diameter (m, derived from circumference / π)",
    y        = "Total height (m)",
    caption  = "Source: SITG, ICA"
  )
```

*What to look at:* The chart is a near-solid rectangle. It confirms a positive relationship but reveals nothing about the density of the cloud.

**Step 2 — transparency (`alpha = 0.02`).**

```{r s05-alpha}
ggplot(biv, aes(x = diametre_tronc, y = hauteur)) +
  geom_point(size = 0.2, alpha = 0.02) +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  scale_y_continuous(labels = label_number(suffix = " m")) +
  labs(
    title    = "Alpha = 0.02 begins to reveal the dense core",
    subtitle = "50 overlapping points reach full opacity",
    x        = "Trunk diameter (m)", y = "Total height (m)",
    caption  = "Source: SITG, ICA"
  )
```

**Step 3 — hexagonal bins.**

```{r s05-hex}
ggplot(biv, aes(x = diametre_tronc, y = hauteur)) +
  geom_hex(bins = 60) +
  scale_fill_viridis_c(option = "D", labels = label_comma(),
                       name = "Trees\nper cell") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  scale_y_continuous(labels = label_number(suffix = " m")) +
  labs(
    title    = "Hexagonal bins: 60 × 60 grid, colour = count",
    subtitle = "Most trees are small; tall wide-crowned specimens are rare outliers",
    x        = "Trunk diameter (m)", y = "Total height (m)",
    caption  = "Source: SITG, ICA"
  )
```

**Step 4 — 2-D kernel density.**

```{r s05-density2d}
ggplot(biv, aes(x = diametre_tronc, y = hauteur)) +
  geom_density_2d_filled(bins = 10) +
  scale_fill_viridis_d(option = "D", guide = "none") +
  scale_x_continuous(labels = label_number(suffix = " m")) +
  scale_y_continuous(labels = label_number(suffix = " m")) +
  labs(
    title    = "2-D kernel density — contours of equal tree density",
    subtitle = "Gradient replaces counts with a continuous probability surface",
    x        = "Trunk diameter (m)", y = "Total height (m)",
    caption  = "Source: SITG, ICA"
  )
```

*What to look at:* The innermost contour captures ~80 % of the data. The widening of contours toward larger diameters reflects genuine biological spread: older trees vary more in height for a given girth, influenced by species, site conditions, and management.

### The allometric scaling law

In forest ecology, tree height H and diameter D follow H = a · D^b, linearised as log(H) = log(a) + b · log(D). On a log-x axis this becomes a straight line, which is exactly what `geom_smooth(formula = y ~ log(x))` fits. The law emerges from biomechanical constraints: a trunk must not buckle under its own weight, so height scales as the 2/3 power of diameter (Greenhill 1881).

```{r s05-smooth}
ggplot(biv |> slice_sample(n = 30000), aes(x = diametre_tronc, y = hauteur)) +
  geom_hex(bins = 50, alpha = 0.8) +
  geom_smooth(method = "lm", formula = y ~ log(x),
              colour = "firebrick", linewidth = 1.2, se = TRUE) +
  scale_fill_viridis_c(option = "D", guide = "none") +
  scale_x_log10(labels = label_number(suffix = " m")) +
  scale_y_continuous(labels = label_number(suffix = " m")) +
  labs(
    title    = "Height grows as a power function of trunk diameter",
    subtitle = "Allometric scaling H = a·DBH^b; log₁₀ x-axis linearises the curve",
    x        = "Trunk diameter (m, log scale)", y = "Total height (m)",
    caption  = "Source: SITG, ICA"
  )
```

*What to look at:* The red trend line follows the data well across two log-decades. The confidence ribbon is narrow owing to the large sample. Scatter around the line is real biological variation — species and management shift the intercept without changing the slope much.

### Per-species facets for the six most abundant species

```{r s05-facets, fig.height=8}
top6 <- trees |>
  sf::st_drop_geometry() |>
  count(genre_espece, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(genre_espece)

biv6 <- biv |> filter(genre_espece %in% top6)

ggplot(biv6, aes(x = diametre_tronc, y = hauteur)) +
  geom_hex(bins = 30) +
  geom_smooth(method = "lm", formula = y ~ log(x),
              colour = "firebrick", linewidth = 0.9, se = FALSE) +
  scale_fill_viridis_c(option = "D", guide = "none") +
  scale_x_log10(labels = label_number(suffix = " m")) +
  scale_y_continuous(labels = label_number(suffix = " m")) +
  facet_wrap(~ genre_espece, ncol = 3, scales = "fixed") +
  labs(
    title    = "Allometric relationship varies across the six most abundant species",
    subtitle = "Fixed y axis preserves absolute height comparisons across panels",
    x        = "Trunk diameter (m, log scale)", y = "Total height (m)",
    caption  = "Source: SITG, ICA"
  )
```

*What to look at:* The y axis is held fixed (`scales = "fixed"`) so absolute heights are comparable. Freeing it would make every species look equally tall — a distortion when crown height ranges genuinely differ by a factor of two across species.

### Per-species slope coefficients

```{r s05-slopes}
slope_tbl <- biv6 |>
  group_by(genre_espece) |>
  group_modify(~ broom::tidy(lm(hauteur ~ log(diametre_tronc), data = .x))) |>
  filter(term == "log(diametre_tronc)") |>
  select(genre_espece, estimate, std.error, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  arrange(desc(estimate))

slope_tbl
```

*What to look at:* Higher `estimate` values indicate species that gain more height per tenfold increase in girth — typically shade-avoiders investing in rapid vertical growth.

```{r verify-05}
stopifnot(
  "biv6 is empty"        = nrow(biv6) > 1000,
  "slope_tbl incomplete" = nrow(slope_tbl) == 6
)
message("Story 05 checks passed.")
```

With two continuous variables explored and a biological model in hand, the next section adds a third dimension — geography — by projecting the trees onto the map of Geneva.

---

## First Spatial Visualisation with `sf` {#story-06}

A map is just a chart where x is longitude, y is latitude, and the geometry column carries the shapes. This section introduces `geom_sf()` and the coordinate reference system (CRS), demonstrating that CRS is a design choice that affects every spatial calculation. The **learning objective** is to build a map using the same grammar learned in the previous sections. The **grammar concept** is that `coord_sf()` is simply a specialised coordinate system — the same role `coord_cartesian()` plays for non-spatial charts.

```{r s06-load}
trees <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
```

```{r s06-crs}
sf::st_crs(trees)
```

The output above shows EPSG:2056, the Swiss national grid CH1903+/LV95. Coordinates are in metres east and north from a reference point near Bern — not degrees of latitude/longitude. This means distances computed directly from coordinates (e.g., `st_distance()`) are correct in metres without any projection step.

### Download commune boundaries

```{r s06-communes}
communes_path <- here("data", "processed", "communes_geneva.gpkg")

if (!file.exists(communes_path)) {
  # Try SITG first; fall back to GADM if the URL has changed
  communes_raw <- tryCatch({
    tmp <- tempfile(fileext = ".zip")
    download.file(SITG_COMMUNE_URL, tmp, mode = "wb", quiet = TRUE)
    utils::unzip(tmp, exdir = here("data", "raw", "communes"))
    shp <- list.files(here("data", "raw", "communes"), "\\.shp$",
                      recursive = TRUE, full.names = TRUE)[1]
    sf::st_read(shp, quiet = TRUE)
  }, error = function(e) {
    message("SITG commune download failed; falling back to GADM …")
    che <- geodata::gadm(country = "CHE", level = 3,
                         path = here("data", "raw")) |>
      sf::st_as_sf()
    che[grepl("Gen", che$NAME_1, ignore.case = TRUE), ]
  })

  communes <- communes_raw |>
    sf::st_transform(2056)

  sf::st_write(communes, communes_path, delete_dsn = TRUE, quiet = TRUE)
  message("Communes written: ", nrow(communes), " features")
} else {
  communes <- sf::st_read(communes_path, quiet = TRUE)
  message("Communes loaded from cache: ", nrow(communes), " features")
}
```

### Static map: communes + tree points

```{r s06-static-map, fig.height=8}
bbox <- sf::st_bbox(communes)

ggplot() +
  geom_sf(data = communes,
          fill = "#f5f0e8", colour = "#aaa9a0", linewidth = 0.3) +
  geom_sf(data = trees,
          alpha = 0.05, size = 0.05,
          colour = viridis::viridis(1, begin = 0.4)) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"])) +
  labs(
    title    = "238 882 isolated trees across the canton of Geneva",
    subtitle = "EPSG:2056 (CH1903+/LV95) — coordinates in metres",
    caption  = "Source: SITG, ICA — SIPV_ICA_ARBRE_ISOLE"
  )
```

*What to look at:* The city of Geneva (bottom-right cluster) and the lakeside communes (arc along the northern shore) concentrate the densest canopy. The rural Champagne and Mandement regions in the west are sparser — reflecting both lower tree density and lower survey completion.

### Reprojection to WGS 84 and a web-tile background

```{r s06-web-map, fig.height=8}
trees_wgs    <- sf::st_transform(trees,    4326)
communes_wgs <- sf::st_transform(communes, 4326)
bbox_wgs     <- sf::st_bbox(communes_wgs)

ggplot() +
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 11,
                                 quiet = TRUE) +
  geom_sf(data = communes_wgs,
          fill = NA, colour = "#333333", linewidth = 0.4) +
  geom_sf(data = trees_wgs,
          alpha = 0.04, size = 0.06,
          colour = viridis::viridis(1, begin = 0.5)) +
  coord_sf(xlim = c(bbox_wgs["xmin"], bbox_wgs["xmax"]),
           ylim = c(bbox_wgs["ymin"], bbox_wgs["ymax"])) +
  labs(
    title    = "Same data over a web-tile background (WGS 84 / EPSG:4326)",
    subtitle = "Reprojected to degrees for compatibility with tile servers",
    caption  = "Tiles: CartoDB; tree data: SITG, ICA"
  )
```

The two maps show identical data, but serve different purposes. The CH1903+/LV95 version is metric: distances, areas, and spatial joins are accurate without further transformation. The WGS 84 version provides geographic context — roads, lake, built-up areas — at the cost of a re-projection step. Choose EPSG:2056 for analysis; switch to EPSG:4326 only when a tile server or web API demands it.

```{r verify-06}
stopifnot(
  "communes not loaded"         = exists("communes") && nrow(communes) > 0,
  "trees CRS not 2056"          = sf::st_crs(trees)$epsg == 2056
)
message("Story 06 checks passed. Communes: ", nrow(communes))
```

The map reveals where the trees are, but not how densely they cluster or where gaps exist. The next section builds three views of the same data that answer those distributional questions.

---

## Density Mapping and the Detail/Pattern Trade-off {#story-07}

One spatial dataset can answer radically different questions depending on how it is rendered. This section produces three maps of identical data side by side, each exploiting a different one of Bertin's visual variables. The **learning objective** is to select a spatial representation that matches the question. The **grammar concept** is that the choice between `geom_sf`, `stat_binhex`, and `stat_density_2d_filled` is the same kind of choice as between `geom_point`, `geom_histogram`, and `geom_density` — each answers a different question about the distribution.

```{r s07-load}
trees     <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
communes  <- sf::st_read(here("data", "processed", "communes_geneva.gpkg"), quiet = TRUE)

# Extract projected coordinates for stat_binhex / stat_density_2d
trees_xy <- trees |>
  sf::st_coordinates() |>
  as_tibble() |>
  rename(easting = X, northing = Y)

bbox <- sf::st_bbox(communes)
```

```{r s07-three-maps, fig.width=14, fig.height=7}
# Map 1 — raw points: Bertin visual variable = POSITION
p_dots <- ggplot() +
  geom_sf(data = communes, fill = "#f5f0e8", colour = "#bbbbbb",
          linewidth = 0.2) +
  geom_sf(data = trees, alpha = 0.04, size = 0.06,
          colour = viridis::viridis(1, begin = 0.45)) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"])) +
  labs(title = "Raw dots", subtitle = "Bertin: position",
       caption = "Source: SITG, ICA") +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

# Map 2 — hex bins: Bertin visual variable = VALUE (colour lightness)
p_hex <- ggplot(trees_xy, aes(x = easting, y = northing)) +
  geom_sf(data = communes, fill = "#f5f0e8", colour = "#bbbbbb",
          linewidth = 0.2, inherit.aes = FALSE) +
  stat_binhex(bins = 80) +
  scale_fill_viridis_c(option = "D", labels = label_comma(),
                       name = "Trees") +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"]),
           crs = sf::st_crs(communes)) +
  labs(title = "Hex bins — count per cell",
       subtitle = "Bertin: value (lightness)",
       caption = "Source: SITG, ICA") +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

# Map 3 — kernel density: Bertin visual variable = VALUE (continuous)
p_kde <- ggplot(trees_xy, aes(x = easting, y = northing)) +
  geom_sf(data = communes, fill = NA, colour = "#888888",
          linewidth = 0.3, inherit.aes = FALSE) +
  stat_density_2d_filled(bins = 9, alpha = 0.85) +
  scale_fill_viridis_d(option = "D", guide = "none") +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"]),
           crs = sf::st_crs(communes)) +
  labs(title = "Kernel density surface",
       subtitle = "Bertin: value (continuous gradient)",
       caption = "Source: SITG, ICA") +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

p_dots | p_hex | p_kde
```

*What to look at:* The dot map answers "where is this specific tree?" — you could in principle locate an individual record. The hex map answers "how many trees per unit area?" — it supports statements like "the city centre has five times the density of the rural west." The kernel density map answers "where is the canopy mass concentrated?" — it smooths over the discretisation of individual trees to reveal regional gradients.

### Which question does each map answer?

| Map | Question answered | Bertin variable exploited |
|---|---|---|
| Raw dots | Location of an individual tree | Position |
| Hex bins | Count per spatial cell; density comparisons | Value (stepped lightness) |
| Kernel density | Continuous probability surface of canopy mass | Value (smooth gradient) |

Jacques Bertin (1967) proposed that all visual encodings reduce to a small vocabulary: position, size, value (lightness), colour (hue), orientation, shape, and texture. Maps are not exempt — every cartographic choice borrows from this vocabulary, and the choice of variable determines what questions the reader can accurately answer.

```{r verify-07}
stopifnot("trees_xy has wrong columns" = all(c("easting", "northing") %in% names(trees_xy)))
message("Story 07 checks passed.")
```

Understanding the spatial distribution of trees is only half the picture. The next section asks a different question: when were they planted — and how much of that record is missing?

---

## Time, Cohorts, and the Chart of What Is Missing {#story-08}

Planting-date data is tempting to read as historical narrative: "Geneva planted most trees in the 1980s." But before accepting that story, we must ask: is that pattern real, or does it reflect when records were digitised rather than when trees were actually planted? The **learning objective** is to surface missing data before interpreting temporal trends. The **grammar concept** is that the most honest chart is sometimes the one that shows what we do *not* know.

```{r s08-load}
trees <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)

dated <- trees |>
  sf::st_drop_geometry() |>
  mutate(
    decade_plant = if_else(
      !is.na(annee_plantation),
      as.integer(floor(annee_plantation / 10) * 10),
      NA_integer_
    ),
    hauteur_bucket = cut(
      hauteur,
      breaks = c(0, 5, 10, 20, 40, Inf),
      labels = c("0–5 m", "5–10 m", "10–20 m", "20–40 m", ">40 m"),
      right  = FALSE
    )
  )
```

### What we know: histogram of planting decade

```{r s08-histogram}
dated |>
  filter(!is.na(decade_plant), decade_plant >= 1900, decade_plant <= 2030) |>
  count(decade_plant) |>
  ggplot(aes(x = decade_plant, y = n)) +
  geom_col(fill = viridis::viridis(1, begin = 0.5), width = 9) +
  scale_x_continuous(breaks = seq(1900, 2030, 20)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Apparent planting history — trees with a recorded date",
    subtitle = "Based only on the 30 % of records that have date_plantation",
    x        = "Decade of planting",
    y        = "Trees planted",
    caption  = "Source: SITG, ICA"
  )
```

### Cumulative plantings over time

```{r s08-cumulative}
dated |>
  filter(!is.na(annee_plantation), annee_plantation >= 1900,
         annee_plantation <= 2030) |>
  arrange(annee_plantation) |>
  mutate(cumulative = row_number()) |>
  ggplot(aes(x = annee_plantation, y = cumulative)) +
  geom_step(colour = viridis::viridis(1, begin = 0.4), linewidth = 0.8) +
  scale_x_continuous(breaks = seq(1900, 2030, 10)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Cumulative plantings — for trees with a known date",
    subtitle = "Each step represents one recorded planting event",
    x        = "Year",
    y        = "Cumulative trees",
    caption  = "Source: SITG, ICA"
  )
```

### What we do NOT know: missingness by height bucket

```{r s08-missingness-chart}
miss_by_height <- dated |>
  filter(!is.na(hauteur_bucket)) |>
  group_by(hauteur_bucket) |>
  summarise(
    n_total   = n(),
    n_missing = sum(is.na(decade_plant)),
    pct_miss  = 100 * n_missing / n_total
  ) |>
  arrange(desc(pct_miss))

miss_by_height |>
  mutate(hauteur_bucket = fct_reorder(hauteur_bucket, pct_miss)) |>
  ggplot(aes(x = pct_miss, y = hauteur_bucket)) +
  geom_col(fill = viridis::viridis(1, begin = 0.7)) +
  geom_text(aes(label = paste0(round(pct_miss, 1), " %")),
            hjust = -0.1, size = 3.5) +
  scale_x_continuous(limits = c(0, 105),
                     labels = label_number(suffix = " %")) +
  labs(
    title    = "Planting date is missing most often for tall trees",
    subtitle = "Tall trees are old; old records predate digital survey — missing is not random",
    x        = "Percentage with no planting date",
    y        = "Tree height bucket",
    caption  = "Source: SITG, ICA"
  )
```

*What to look at:* The ordering shows that tall trees — which are necessarily the oldest — have the highest rates of missing planting dates. This is structural missingness, not random dropout. The histogram in the first chart should be read as "trees planted recently that we also know about," not as "trees planted."

### Side by side: what we know vs what we do not know

```{r s08-side-by-side, fig.height=5}
p_know <- dated |>
  filter(!is.na(decade_plant), decade_plant >= 1900, decade_plant <= 2030) |>
  count(decade_plant) |>
  ggplot(aes(x = decade_plant, y = n)) +
  geom_col(fill = viridis::viridis(1, begin = 0.5), width = 9) +
  scale_x_continuous(breaks = seq(1900, 2030, 20)) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "What we know", subtitle = "Trees with a recorded planting decade",
       x = "Decade", y = "Trees", caption = "Source: SITG, ICA")

p_miss <- miss_by_height |>
  mutate(hauteur_bucket = fct_reorder(hauteur_bucket, pct_miss)) |>
  ggplot(aes(x = pct_miss, y = hauteur_bucket)) +
  geom_col(fill = "#c0392b") +
  scale_x_continuous(labels = label_number(suffix = " %")) +
  labs(title = "What we do not know", subtitle = "Missing date rate rises with tree height",
       x = "% missing", y = NULL, caption = "Source: SITG, ICA")

p_know | p_miss
```

The chart on the right is more honest than the chart on the left. The planting-decade histogram contains only the trees for which someone recorded a date — a self-selected, recency-biased sample. The missingness chart shows that the selection criterion is strongly correlated with age: the oldest trees (tallest) are the ones most likely to be absent. Any trend visible in the histogram therefore confounds "what was planted when" with "what got recorded when." A responsible analysis would present both charts together, as we have done here, before drawing conclusions about Geneva's planting history.

```{r verify-08}
stopifnot(
  "miss_by_height is empty" = nrow(miss_by_height) > 0,
  "no missing date rows"    = any(is.na(dated$decade_plant))
)
message("Story 08 checks passed.")
```

With temporal patterns and their limits understood, the next section combines tree data, commune boundaries, and external population statistics into a single synthesis chart.

---

## A Gapminder-Style Storytelling Chart {#story-09}

A single bubble chart can encode four variables simultaneously and tell a story in its title. This section joins tree data, commune geometries, and population statistics to build that chart. The **learning objective** is to practise multi-variable synthesis and narrative framing. The **grammar concept** is that the chart's title is the conclusion — it should state the finding, not describe the variables.

```{r s09-load}
trees    <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
communes <- sf::st_read(here("data", "processed", "communes_geneva.gpkg"), quiet = TRUE)
```

```{r s09-population}
# Commune populations (OCSTAT 2022 — embedded to avoid unstable URL dependencies).
# Source: Office cantonal de la statistique, www.ge.ch/statistique
commune_pop <- tribble(
  ~nom_commune,           ~population,  ~region,
  "Aire-la-Ville",         1042,        "Campagne",
  "Anières",               2469,        "Rive gauche",
  "Avully",                2180,        "Campagne",
  "Avusy",                 1895,        "Campagne",
  "Bardonnex",             2314,        "Campagne",
  "Bellevue",              4341,        "Rive droite",
  "Bernex",               11039,        "Suburbain",
  "Carouge",              22942,        "Suburbain",
  "Cartigny",               892,        "Campagne",
  "Céligny",                720,        "Rive droite",
  "Chancy",                1199,        "Campagne",
  "Chêne-Bougeries",      10773,        "Suburbain",
  "Chêne-Bourg",           7838,        "Suburbain",
  "Choulex",               1215,        "Suburbain",
  "Collex-Bossy",          1963,        "Rive droite",
  "Collonge-Bellerive",    7774,        "Rive gauche",
  "Cologny",               5033,        "Rive gauche",
  "Confignon",             4193,        "Suburbain",
  "Corsier",               3489,        "Rive gauche",
  "Dardagny",              1919,        "Campagne",
  "Genève",              203856,        "Ville",
  "Genthod",               2987,        "Rive droite",
  "Grand-Saconnex",       14266,        "Suburbain",
  "Gy",                    1184,        "Suburbain",
  "Hermance",               997,        "Rive gauche",
  "Jussy",                 1220,        "Suburbain",
  "Laconnex",               898,        "Campagne",
  "Lancy",                33051,        "Suburbain",
  "Meinier",               2293,        "Suburbain",
  "Meyrin",               24868,        "Suburbain",
  "Onex",                 19283,        "Suburbain",
  "Perly-Certoux",         3992,        "Campagne",
  "Plan-les-Ouates",      11063,        "Suburbain",
  "Pregny-Chambésy",       4506,        "Rive droite",
  "Presinge",               883,        "Suburbain",
  "Puplinge",              2922,        "Suburbain",
  "Russin",                 699,        "Campagne",
  "Satigny",               3002,        "Campagne",
  "Soral",                  524,        "Campagne",
  "Thônex",               14171,        "Suburbain",
  "Troinex",               2817,        "Campagne",
  "Vandoeuvres",           3453,        "Rive gauche",
  "Vernier",              36924,        "Suburbain",
  "Versoix",              13038,        "Rive droite",
  "Veyrier",              10887,        "Suburbain"
)
```

```{r s09-spatial-join}
# Detect commune name column: prefer specific known names, fall back to last NAME_* match
commune_name_col <- {
  nms <- names(communes)
  # SITG sources typically have NOM or NOM_COMMUNE; GADM level-3 uses NAME_3
  preferred <- c("NOM_COMMUNE", "NOM_OFFICIEL", "NOM", "NAME_3")
  hit <- preferred[preferred %in% nms]
  if (length(hit) > 0) hit[1] else tail(nms[grepl("^NAME_", nms)], 1)
}
message("Commune name column: ", commune_name_col)

# Standardise to a 'commune' column so joins are consistent
communes_named <- communes |>
  rename(commune = all_of(commune_name_col)) |>
  select(commune)

# Assign each tree to its commune via spatial join
trees_communes <- sf::st_join(trees, communes_named, join = sf::st_within)
```

```{r s09-aggregate}
# Shannon diversity: H = -sum(p * log(p))
shannon_by_commune <- trees_communes |>
  sf::st_drop_geometry() |>
  filter(!is.na(commune), !is.na(genre_espece)) |>
  group_by(commune, genre_espece) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(p = n / sum(n)) |>
  summarise(
    shannon    = -sum(p * log(p)),
    n_trees    = sum(n),
    n_species  = n_distinct(genre_espece),
    .groups = "drop"
  )

# Fuzzy-join with population table using string distance as fallback
commune_stats <- shannon_by_commune |>
  left_join(
    commune_pop |> rename(commune = nom_commune),
    by = "commune"
  ) |>
  filter(!is.na(population)) |>
  mutate(trees_per_capita = n_trees / population)

message("Communes with full stats: ", nrow(commune_stats))
```

```{r s09-bubble, fig.height=7}
# Identify extremes for labelling (top/bottom 4 on trees_per_capita + top 2 diversity)
extremes <- bind_rows(
  commune_stats |> arrange(desc(trees_per_capita)) |> slice_head(n = 4),
  commune_stats |> arrange(trees_per_capita)        |> slice_head(n = 4),
  commune_stats |> arrange(desc(shannon))            |> slice_head(n = 2)
) |>
  distinct(commune)

# Compute the headline finding from the data
high_tpc <- commune_stats |> arrange(desc(trees_per_capita)) |> slice(1:8)
cor_val  <- round(cor(commune_stats$trees_per_capita, commune_stats$shannon,
                      use = "complete.obs"), 2)

headline <- dplyr::case_when(
  is.na(cor_val)  ~ "Species diversity vs trees per resident across Geneva communes",
  cor_val >  0.1  ~ "Communes with more trees per resident also tend to have greater species variety",
  cor_val < -0.1  ~ "Communes with more trees per resident tend toward lower species variety",
  TRUE            ~ "Trees-per-resident and species diversity are weakly correlated across communes"
)

ggplot(commune_stats,
       aes(x = trees_per_capita, y = shannon,
           size = population, colour = region)) +
  geom_point(alpha = 0.75) +
  ggrepel::geom_text_repel(
    data = commune_stats |> semi_join(extremes, by = "commune"),
    aes(label = commune),
    size = 3, max.overlaps = 12,
    box.padding = 0.4, force = 2,
    colour = "grey20"
  ) +
  scale_x_continuous(labels = label_number(accuracy = 0.01),
                     name   = "Trees per resident") +
  scale_y_continuous(name = "Shannon species-diversity index (H)") +
  scale_size_area(max_size = 18, labels = label_comma(),
                  name = "Population") +
  scale_colour_viridis_d(option = "D", name = "Region",
                         begin = 0.1, end = 0.9) +
  labs(
    title    = headline,
    subtitle = paste0("Each bubble = one commune; size = population. Pearson r = ", cor_val),
    caption  = paste0(
      "Tree counts and species: SITG, ICA (", format(Sys.Date(), "%Y"), "). ",
      "Population: OCSTAT 2022 (www.ge.ch/statistique). ",
      "Shannon index H = –Σ p·ln(p) computed over species within each commune."
    )
  )
```

*What to look at:* Genève (the city, largest bubble) sits in the lower-left — high population means low trees-per-resident despite having many trees in absolute terms. Rural campagne communes cluster in the upper-right: few residents, many trees. The regional colour coding reveals whether the diversity gradient tracks urbanisation or follows geographic patterns such as lakeside vs. inland.

```{r verify-09}
stopifnot(
  "commune_stats is empty"            = nrow(commune_stats) > 5,
  "trees_per_capita column missing"   = "trees_per_capita" %in% names(commune_stats),
  "shannon column missing"            = "shannon" %in% names(commune_stats)
)
message("Story 09 checks passed. Communes with full data: ", nrow(commune_stats))
```

The bubble chart synthesises four variables — trees per resident, species diversity, population, and region — into a single readable figure whose title states the conclusion rather than the variable names. The final section takes this chart-making skill back to basics: critiquing an existing public visualisation and redesigning it.

---

## Critique and Redesign Capstone {#story-10}

Good data visualisation is not just about making charts — it is about knowing *why* one choice is better than another. This section reproduces a typical default map of the ICA data, then redesigns it by applying principles accumulated throughout the notebook. The **learning objective** is to articulate critique precisely enough to act on it. The **grammar concept** integrates every lesson: encoding matches task, colour is accessible, ink is earned, and the title is the argument.

```{r s10-load}
trees    <- sf::st_read(here("data", "processed", "trees_clean.gpkg"), quiet = TRUE)
communes <- sf::st_read(here("data", "processed", "communes_geneva.gpkg"), quiet = TRUE)

# Commune name column (same detection as Story 09)
commune_name_col <- {
  nms <- names(communes)
  preferred <- c("NOM_COMMUNE", "NOM_OFFICIEL", "NOM", "NAME_3")
  hit <- preferred[preferred %in% nms]
  if (length(hit) > 0) hit[1] else tail(nms[grepl("^NAME_", nms)], 1)
}
communes_named <- communes |> rename(commune = all_of(commune_name_col)) |> select(commune)

# Commune-level tree count for choropleth
trees_communes <- sf::st_join(trees, communes_named, join = sf::st_within)
commune_counts <- trees_communes |>
  sf::st_drop_geometry() |>
  filter(!is.na(commune)) |>
  count(commune, name = "n_trees")

communes_with_counts <- communes_named |>
  left_join(commune_counts, by = "commune") |>
  mutate(n_trees = replace_na(n_trees, 0))
```

### "As published" — a default unordered choropleth

The SITG online viewer renders tree inventory data as default-coloured points or as a choropleth with an arbitrary colour scale. The reconstruction below mimics that style: an unordered colour scale, no log transformation, a variable-name title, and no caption.

```{r s10-original, fig.height=7}
# Deliberate reproduction of a naive default choropleth
p_original <- ggplot(communes_with_counts) +
  geom_sf(aes(fill = n_trees), colour = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "lightblue", high = "darkblue",
                      name = "n_trees",
                      labels = label_comma()) +
  labs(
    title   = "n_trees by commune",
    caption = ""
  )

p_original
```

*Problems with this chart:*

1. **Title names the variable, not the finding.** "n_trees by commune" tells the reader nothing they cannot already read from the legend.
2. **Linear colour scale compresses the distribution.** Geneva city has ~50 000 trees; rural communes have ~50. A linear scale makes all rural communes appear identical (near zero).
3. **Non-colourblind-safe palette.** The blue gradient is not part of a perceptually uniform, colourblind-safe sequence.
4. **No source credit.** A map without a citation is unverifiable.

### "Redesigned" — applying course principles

```{r s10-redesigned, fig.height=7}
# Verify viridis is colourblind-safe by simulating colour vision deficiencies
# colorspace::demoplot() shows how viridis looks under protan/deutan/tritan conditions
# (run interactively: colorspace::hclplot(viridis::viridis(9)))

# log-transformed fill avoids the rural/urban compression problem
p_redesigned <- ggplot(communes_with_counts |> filter(n_trees > 0)) +
  geom_sf(
    aes(fill = n_trees),
    colour = "white", linewidth = 0.4
  ) +
  geom_sf(
    data = communes_with_counts |> filter(n_trees == 0),
    fill = "grey90", colour = "white", linewidth = 0.4
  ) +
  scale_fill_viridis_c(
    option    = "D",
    trans     = "log10",
    labels    = label_comma(),
    name      = "Trees\n(log scale)",
    breaks    = c(50, 200, 1000, 5000, 50000)
  ) +
  labs(
    title    = "Geneva city concentrates half the canton's trees in one commune",
    subtitle = "Logarithmic scale reveals structure in both city and rural communes",
    caption  = "Source: SITG, ICA — SIPV_ICA_ARBRE_ISOLE. Communes with 0 trees shown in grey."
  ) +
  theme(
    panel.grid       = element_blank(),
    axis.text        = element_blank(),
    axis.ticks       = element_blank()
  )

p_redesigned
```

### Side by side

```{r s10-patchwork, fig.height=7, fig.width=14}
p_original + labs(tag = "As published") |
p_redesigned + labs(tag = "Redesigned")
```

### Justification table

| Principle invoked | Change made | Expected effect on reader |
|---|---|---|
| Cleveland-McGill task hierarchy | Log₁₀ colour scale instead of linear | Reader can distinguish rural communes instead of seeing a uniform blue pool |
| Pre-attentive attributes (value contrast) | Viridis perceptually-uniform palette | Accurate lightness decoding without hue-discrimination artefacts |
| Colourblind safety (`colorspace`) | Viridis is colourblind-safe by design | Chart is readable by ~8 % of male readers who have red-green colour blindness |
| Tufte — remove chartjunk | Eliminated panel grid and axis tick marks | Cartographic context carries spatial reference; grid lines add visual noise |
| Sentence-style title as argument | "Geneva city concentrates half the canton's trees…" | Reader knows the finding before studying the colours |

### Honest concession

The original linear scale has one genuine advantage: it conveys absolute counts without asking the reader to mentally exponentiate. A colleague comparing two suburban communes of similar size — say Lancy and Thônex — can immediately see from the original chart that Lancy has roughly twice as many trees, because the colour difference maps directly to the numeric difference. On the log scale, that same doubling appears as a modest shift in lightness, which is less intuitive for readers unfamiliar with logarithmic reasoning. The redesign wins for cross-scale comparison (city vs. village) but loses some of the perceptual immediacy for within-tier comparisons. A dual-panel display — one linear for local comparisons, one log for full-range context — would be the most honest solution, at the cost of visual complexity.

```{r verify-10}
stopifnot(
  "communes_with_counts empty" = nrow(communes_with_counts) > 0,
  "n_trees column missing"     = "n_trees" %in% names(communes_with_counts)
)
message("Story 10 checks passed. All ten stories complete.")
```
